---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 5: Exploratory Data Analysis"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 24, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
    - ggrepel
    - patchwork
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
smaller <- diamonds |> filter(carat < 3)
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
smaller <- diamonds |> filter(carat < 3)
```

## The EDA loop

EDA is not a procedure with a checklist. It's a cycle you run until the
questions stop being interesting:

1. **Generate questions** about your data
2. **Search for answers** by visualizing, transforming, and modelling
3. **Refine** your questions, or generate new ones from what you found

Most of what you produce during EDA gets thrown away, and that's okay!
 — the output is *understanding*, not plots.


## Why it's not optional

::: {.callout-note title="Important note"}
Dead ends are expected and not failure. 
:::

Even when the research question arrives fully formed, you still have to know whether the data can answer it.


::: {.callout-caution title="Data cleaning"}

A *special case* of EDA: you're asking whether the data meets your expectations, and using the same tools like visualization, transformation, modelling to find out.

:::

## Setup

```{r}
#| eval: false
library(tidyverse)
```

Two guiding questions carry most of the weight:

1. What **variation** occurs *within* my variables?
2. What **covariation** occurs *between* my variables?

::: {.callout-warning title="Remember from the Financial Wellness Data?}

We had three clusters for financial wellness: stressed, coping , and comfortable. What were the variation and covariation discussed in that example?

:::


# Variation

## Variation

::: {.callout-note title="Variation"}
The tendency of a variable's values to change from one measurement to the
next.
:::

- Even a perfectly constant quantity will read slightly differently each time you measure it. That's measurement noise. 

- Real variables also differ between subjects and over time on top of that noise, and the combination gives every variable its own characteristic shape. 

- Some cluster tightly, some spread wide, some have several common values.

- You can't see that shape in a long column of numbers. **Plotting the
distribution is how you see it.**


## The Diamond Data {.smaller}

::: {.callout-note title="The `diamonds` variables"}
| Variable | Meaning |
|---|---|
| `price` | price in US dollars |
| `carat` | weight of the diamond |
| `cut` | quality of the cut: Fair, Good, Very Good, Premium, Ideal |
| `color` | diamond color, from J (worst) to D (best) |
| `clarity` | how clear the diamond is: I1 (worst) to IF (best) |
| `x`, `y`, `z` | length, width, depth in mm |
| `depth` | total depth percentage |
| `table` | width of the top of the diamond relative to widest point |
:::

## A first look

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

## Changing the Binwidth 

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Same data. A much narrower bin reveals structure the first plot hid entirely.

## Typical values: what to ask

Tall bars are common values, short bars rare ones, gaps are values never observed. Turn that into questions:

- Which values are most common? Why?
- Which are rare? Does that match what you expected?
- Are there unexpected patterns? What could explain them?

For the carat histogram: why the spikes at whole carats and common fractions? And why more diamonds just *above* each spike than just below?

::: {.notes}
The answer involves both cutting practice and pricing incentives — a good
illustration that some EDA questions need domain knowledge, not more code.
:::

## Clusters mean subgroups

When a distribution separates into clumps, you can ask yourself:

- How are observations *within* a cluster alike?
- How do clusters differ from each other?
- What explains the split?
- Could the clusters be an artifact rather than real structure?

Many of these push you toward a relationship *between* variables.

## Unusual values hide in plain sight

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5)
```

The only hint of anything odd is how wide the x-axis is.

## Zoom in on the y-axis

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```



## Zooming In

Now three unusual regions show up: near 0, around 30, and around 60.

::: {.callout-note title="`coord_cartesian()`"}
The bars are still computed from *all* the data, so you can safely zoom the y-axis to see rare bins without changing what got counted. 

`ylim()`/`xlim()` instead **discard** any data outside the limits before the stat runs, which can silently change your bar heights. 

When you just want a closer look, reach for `coord_cartesian()`.
:::

## Pull the outliers out

```{r}
unusual <- diamonds |>
  filter(y < 3 | y > 20) |>
  select(price, x, y, z) |>
  arrange(y)

unusual
```

## What we found

`y` is a dimension in millimeters. A diamond cannot be 0 mm wide, so those
rows are **missing data encoded as zero**.

  — something no search for `NA` would ever have surfaced.

The 32 mm and 59 mm entries are implausible too!

::: {.callout-important}
This is the core payoff of EDA: visualizing the data surfaced a data-quality problem that a search for `NA` never would have. Always sanity-check units and physically-plausible ranges, not just missingness flags.
:::

::: {.callout-caution}
Run your analysis with and without the outliers. If it barely moves and you can't explain them, dropping them is defensible. If it moves a lot, find the cause and **disclose the removal** in your write-up — don't drop values without justification just because they're inconvenient.
:::

## Two ways to handle them

```{r}
#| eval: false
# Option 1: drop the whole row — usually too blunt
diamonds2 <- diamonds |> filter(between(y, 3, 20))

# Option 2: replace the bad value with NA — usually better
diamonds2 <- diamonds |>
  mutate(y = if_else(y < 3 | y > 20, NA, y))
```

One invalid measurement doesn't invalidate the rest of the observation. Apply
option 1 across every variable in messy data and you may have nothing left.

## Missing values in plots

ggplot2 won't guess where to put an `NA`, so it drops them and warns you. Silence the warning with `na.rm = TRUE` (only once you've understood why they're missing).

```{r}
#| fig-height: 2.9
#| warning: true
diamonds2 <- diamonds |> mutate(y = if_else(y < 3 | y > 20, NA, y))

ggplot(diamonds2, aes(x = x, y = y)) +
  geom_point()
```

## Missingness can be the signal

In `nycflights13::flights`, a missing `dep_time` means the flight was cancelled. So compare scheduled departure times for cancelled and non-cancelled flights:

```{r}
#| fig-height: 2.8
nycflights13::flights |>
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min  = sched_dep_time %% 100,
    sched_dep_time = sched_hour + (sched_min / 60)
  ) |>
  ggplot(aes(x = sched_dep_time)) +
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4)
```


## The comparison isn't fair yet

::: {.callout-tip}
Missing values aren't always noise to drop. Sometimes *why* a value is missing is itself the most informative variable in the dataset — `is.na(dep_time)` here is doing more analytical work than `dep_time` itself.
:::

Cancelled flights are vastly outnumbered, so the two curves aren't comparable on a count scale.

That's a covariation problem, and the next section is about fixing exactly this kind of comparison.

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Explore the distributions of `x`, `y`, and `z` in `diamonds`. Which is length, which width, which depth?
2. Explore `price`. Try a wide range of binwidths. Anything surprising?
3. How many diamonds are 0.99 carat vs. exactly 1 carat? Why the gap?
:::

```{webr}
#| exercise: variation

ggplot(diamonds, aes(x = x)) +
  geom_histogram(binwidth = 0.25)

## Q3
# diamonds |> filter(carat %in% c(0.99, 1.00)) |> count(carat)
```

# Covariation

## Covariation

If variation describes behavior *within* a variable, **covariation** describes behavior *between* variables: the tendency for two or more variables to move together.

The strategy depends on the variable types:

- categorical × numerical
- categorical × categorical
- numerical × numerical

## Categorical × numerical:

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = price)) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

::: {.callout-caution title="The Count Problem"}
Group sizes differ so much that shape differences are invisible. Do you think each `cut` share the same pattern?
:::


## Switch the y-axis to density

**Density** is the count standardized so the area under each curve is 1.
Since it isn't a column in the data, `after_stat()` fetches it from the stat.

```{r}
#| fig-height: 3
ggplot(diamonds, aes(x = price, y = after_stat(density))) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

## Frequency polygons vs Boxplots

Fair diamonds (the *lowest* quality cut) appear to have the highest average price.


```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```


## Reordering 

`cut` is an ordered factor, so its order is meaningful. Most categorical
variables aren't.

::: {.callout-tip title="`fct_reorder(variable, by, fun)`"}
This function sorts factor levels by a summary statistic (e.g. mean/median) of another variable, which almost always makes a categorical-vs-numeric plot easier to read at a glance.
:::


## Example

```{r}
#| fig-height: 3.2
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot()
ggplot(mpg, aes(x = fct_reorder(class, hwy, median), y = hwy)) + geom_boxplot()
```

## Flip when labels are long

Swap the x and y mappings with no extra function needed:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = hwy, y = fct_reorder(class, hwy, median))) +
  geom_boxplot()
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
Recall that there are variables such as `price`, `carat`, `cut`, `color`, `clarity`, `x`, `y`, `z`, `depth`, and `table`. 

Earlier we saw that `Fair` cut diamonds look more expensive on average than `Ideal` cut diamonds, which may be the opposite of what you'd expect. `carat` (the diamond's weight) turns out to be the strongest predictor of `price` in the whole dataset.

1. Plot `price` against `carat`. Does the relationship look strong?
2. Now plot `carat` against `cut`. Is `carat` spread evenly across cut
   qualities, or do `Fair` diamonds skew toward one end?
3. Put it together: why does `Fair` diamonds skewing that way in `carat`
   make them look pricier overall, even though a worse cut should, on its
   own, lower the price?
   
:::

```{webr}
#| exercise: carat_effect

# Step 1: price vs. carat
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/50)

# Step 2: carat vs. cut
ggplot(diamonds, aes(x = cut, y = carat)) +
  geom_boxplot()
```

## Categorical × categorical: counts

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()
```

Circle area shows how many observations fall at each combination.

## Or compute the counts, then tile

```{r}
#| fig-height: 3.2
diamonds |>
  count(color, cut) |>
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
```

::: {.callout-note}
For unordered categories, the **seriation** package can reorder rows and columns to expose structure; **heatmaply** does interactive versions of the same idea for larger tables.
:::

## Numerical × numerical

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point()
```

A positive, roughly exponential relationship, and a **LOT** of overplotting.

## Transparency helps

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point(alpha = 1 / 100)
```

Clusters at 1, 1.5, and 2 carats become visible.

## Bin + Boxplots ?

```{r}
#| fig-height: 3.2
ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

`cut_width(x, width)` slices `x` into equal-width bins. `cut_number()` makes bins with roughly equal *counts* instead.



# Patterns and models

## Interrogating a pattern

A systematic relationship shows up as a pattern. When you spot one, ask:

- Could this be coincidence?
- How would you describe the relationship?
- How strong is it?
- What other variables might be involved?
- Does it hold within subgroups?

## Variation vs. covariation

**Variation** is uncertainty within a single variable. If `price` varies a lot on its own, knowing nothing else, you can't say much about what a given diamond costs. **Covariation** between variables cuts that uncertainty down.

::: {.callout-tip}
If `carat` and `price` covary, then knowing a diamond's `carat` narrows down your guess for its `price` a lot more than guessing blind. The stronger the covariation, the more one variable tells you about the other.
:::


## Modelling

This is the entire logic behind prediction and modelling: a useful predictor is just a variable that covaries with the outcome you care about. A variable that doesn't covary with the outcome is useless as a predictor, no matter how interesting it looks on its own.

::: {.callout-important}
Covariation is not the same as causation. Two variables can covary because one causes the other, because both are driven by a third variable, or by pure coincidence. If the covariation genuinely is causal, you can go a step further than prediction and actually **control** one variable by changing the other. Most covariation you find during EDA is not causal, and confirming causation takes more than a plot.
:::



## Summary

- EDA is a **loop**: question → visualize/transform/model → better questions
- **Variation** within a variable: typical values, clusters, unusual values
- Outliers are sometimes errors, sometimes discoveries. Investigate before deciding; disclose what you removed (**important**)
- **Covariation** between variables: pick the technique by variable type

Next: turning exploratory graphics into ones other people can read.
