---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 3: The Whole Game - Day 2"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 17, 2026"
format: 
  live-revealjs:
    resources:
      - data/raw_survey_2017.csv
      - data/raw_survey_2022.csv
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
---

## Recap: Day 1

- We toured real findings — income, benefits, paycheque-delay difficulty
- We ended with three financial wellness clusters: Stressed, Coping, Comfortable
- Today: how we actually got there — including the messy parts

---

## Before we look at the raw data...

<div style="text-align: center; font-size: 1.3em; margin-top: 1.5em;">
What do you think could go wrong with real, raw survey data?

<br><br>
What kinds of problems might it have that yesterday's polished plots didn't show?
</div>

<!-- Collect answers on screen/board -- don't organize yet, just capture them.
     This list becomes the roadmap for today, and we return to it at the end. -->

---

## The reveal: here's the actual raw file

```{r}
#| eval: false
library(tidyverse)

survey_2017_raw <- read_csv("data/raw_survey_2017.csv")

# Note the skip = 1 -- 2022's file has a stray metadata line above the
# real header (a leftover from the export on the new survey platform)
survey_2022_raw <- read_csv("data/raw_survey_2022.csv", skip = 1)

glimpse(survey_2017_raw)
glimpse(survey_2022_raw)
```

<!-- Let the mess speak for itself. Pause. Validate whatever the class
     predicted a moment ago. -->

---

## First things first: duplicate rows

A few respondents show up more than once (simulating a double-submitted survey).

```{r}
#| eval: false
survey_2017_raw |> count(respondent_id, sort = TRUE) |> filter(n > 1)

survey_2017_raw <- survey_2017_raw |> distinct()
survey_2022_raw <- survey_2022_raw |> distinct()
```

`distinct()` removes exact duplicate rows — simple, but always worth checking
for *before* anything else, since duplicates quietly inflate every count and
mean you compute downstream.

---

## Tidy data: benefits are stuck in wide format

```{r}
#| eval: false
survey_2017_raw |>
  select(respondent_id, starts_with("benefit_")) |>
  head()
```

One respondent, one true preference — but it's spread across six columns.
This needs `pivot_longer()` before we can do anything useful with it.

---

## Tidy data: `pivot_longer()` to the rescue

```{r}
#| eval: false
benefits_2017_long <- survey_2017_raw |>
  pivot_longer(
    cols = starts_with("benefit_"),
    names_to = "benefit_type",
    values_to = "response"
  ) |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(
    benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_,
    .groups = "drop"
  )

survey_2017_clean <- survey_2017_raw |>
  select(-starts_with("benefit_")) |>
  left_join(benefits_2017_long, by = "respondent_id")
```

<!-- Same pattern applies to 2022 -- it just has one fewer benefit_ column,
     since "Educational funding" never existed as an option that year. -->

```{r}
#| eval: false
benefits_2022_long <- survey_2022_raw |>
  pivot_longer(
    cols = starts_with("benefit_"),
    names_to = "benefit_type",
    values_to = "response"
  ) |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(
    benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_,
    .groups = "drop"
  )

survey_2022_clean <- survey_2022_raw |>
  select(-starts_with("benefit_")) |>
  left_join(benefits_2022_long, by = "respondent_id")
```

---

## Strings & expressions: income is a mess

```{r}
#| eval: false
survey_2017_clean |> distinct(income) |> slice_sample(n = 6)
```

Four different formats mixed together: `"85000"`, `"$85,000"`, `"85,000"`,
`"85K"`. Before income can be used in any calculation, it needs to become a
real number.

```{r}
#| eval: false
clean_income <- function(x) {
  x <- str_trim(x)
  is_k <- str_detect(x, "K$")
  numeric_part <- x |>
    str_remove_all("[$,K]") |>
    as.numeric()
  if_else(is_k, numeric_part * 1000, numeric_part)
}

survey_2017_clean <- survey_2017_clean |>
  mutate(income = clean_income(income))

survey_2022_clean <- survey_2022_clean |>
  mutate(income = clean_income(income))
```

---

## Strings & expressions: province spelling

```{r}
#| eval: false
survey_2017_clean |> distinct(province)
```

Same province, several spellings: `"Ontario"`, `"ON"`, `"ontario "`.

```{r}
#| eval: false
standardize_province <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(
    x %in% c("ontario", "on") ~ "Ontario",
    x %in% c("alberta", "ab") ~ "Alberta",
    x %in% c("british columbia", "bc") ~ "British Columbia",
    x %in% c("quebec", "qc") ~ "Quebec",
    TRUE ~ str_to_title(x)
  )
}

survey_2017_clean <- survey_2017_clean |> mutate(province = standardize_province(province))
survey_2022_clean <- survey_2022_clean |> mutate(province = standardize_province(province))
```

---

## Strings & expressions: Yes/No encodings

```{r}
#| eval: false
survey_2017_clean |> distinct(debt_overwhelmed)
```

Three encodings for the same two answers: `"Yes"/"No"`, `"Y"/"N"`, `"yes"/"no"`.

```{r}
#| eval: false
standardize_yn <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(
    x %in% c("yes", "y") ~ "Yes",
    x %in% c("no", "n") ~ "No",
    TRUE ~ NA_character_
  )
}

survey_2017_clean <- survey_2017_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))
survey_2022_clean <- survey_2022_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))
```

---

## Missing data: how much, and where?

```{r}
#| eval: false
survey_2017_clean |>
  summarize(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

---

## Missing data: two honest options

::: {.columns}
::: {.column width="48%"}
**Option 1: Delete**
```{r}
#| eval: false
survey_2017_nomiss <- survey_2017_clean |>
  drop_na(income, savings_pct,
          debt_overwhelmed,
          emergency_fund_likelihood)
```
Simple, but shrinks your sample —
and if missingness isn't random,
you bias what's left.
:::

::: {.column width="48%"}
**Option 2: Impute**
```{r}
#| eval: false
survey_2017_imputed <- survey_2017_clean |>
  mutate(savings_pct = if_else(
    is.na(savings_pct),
    median(savings_pct, na.rm = TRUE),
    savings_pct
  ))
```
Keeps every row, but every imputed
value is a guess — and a bad guess
can quietly distort your results.
:::
:::

<!-- Discussion: which would you choose for savings_pct? For debt_overwhelmed
     (categorical)? Does the answer change per-column? -->

---

## Mismatched response: too many response levels

```{r}
#| eval: false
survey_2017_clean |> distinct(emergency_fund_likelihood)
survey_2022_clean |> distinct(emergency_fund_likelihood)
```

2017 has 7 levels. 2022 already collapsed to 5. Neither matches the other,
and neither is easy to compare across years as-is.

**The compromise:** collapse both down to 3 levels — Difficult, Neither,
Manageable.

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult", "Somewhat difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Somewhat manageable", "Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))

survey_2022_clean <- survey_2022_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))
```

<!-- Worth naming out loud: this is a judgment call. A different compromise
     (e.g. 5 levels, or a different cut point) would give slightly different
     results downstream. -->

---

## Mismatched question: same question, different name

```{r}
#| eval: false
names(survey_2017_clean)
names(survey_2022_clean)
```

`paycheque_delay_difficulty` (2017) and `twoweek_pay_delay_impact` (2022)
ask the same thing — the survey platform changed, the wording changed, the
column name changed. `rename()` fixes this in one line each.

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  rename(pay_delay_response = paycheque_delay_difficulty)

survey_2022_clean <- survey_2022_clean |>
  rename(pay_delay_response = twoweek_pay_delay_impact)
```

---

## Age: numbers in one year, buckets in the other

```{r}
#| eval: false
survey_2017_clean |> select(age) |> head()
survey_2022_clean |> select(age) |> head()
```

2017 gives an exact age. 2022 gives a bracket. To compare years, everyone
needs to speak the same bucket language.

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(age_bucket = case_when(
    age <= 30 ~ "18-30",
    age <= 40 ~ "30-40",
    age <= 50 ~ "40-50",
    age <= 65 ~ "50-65",
    TRUE ~ "65+"
  ))

survey_2022_clean <- survey_2022_clean |>
  rename(age_bucket = age)
```

<!-- Worth a pause: 2022's brackets overlap at the boundaries (is age 30 in
     "18-30" or "30-40"?). Real survey design flaw -- we had to pick a rule
     for 2017 to match, and that choice is itself a judgment call. -->

---

## Removed benefit option: aligning categories

```{r}
#| eval: false
survey_2017_clean |> distinct(benefit_preference)
survey_2022_clean |> distinct(benefit_preference)
```

"Educational funding" existed as a choice in 2017, but the 2022 platform
dropped it. To compare years fairly, we fold anything not offered in both
years into a shared **"Other"** category.

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(benefit_preference = if_else(
    benefit_preference == "benefit_education",
    "benefit_other",
    benefit_preference
  ))
```

---

## Combine years

```{r}
#| eval: false
survey_combined <- bind_rows(survey_2017_clean, survey_2022_clean)

glimpse(survey_combined)
```

Only possible now — column names match (`rename()`), category sets match
(the "Other" recode), and the response scales match (the 3-level collapse).
Every fix so far existed to make this one line possible.

---

## A useful side quest: `separate()`

```{r}
#| eval: false
survey_combined |>
  separate(respondent_id, into = c("survey_year", "respondent_num"), sep = "-") |>
  select(survey_year, respondent_num, everything()) |>
  head()
```

`respondent_id` (`"2017-0001"`) is actually two pieces of information
squeezed into one column. `separate()` is `pivot_longer()`'s cousin for
this specific problem — splitting one messy column into several clean ones.

---

## `pivot_wider()`: the mirror image

```{r}
#| eval: false
survey_combined |>
  count(financial_wellness_cluster, year) |>
  pivot_wider(names_from = year, values_from = n)
```

`pivot_longer()` took wide data and made it long. `pivot_wider()` does the
opposite — useful here for turning a long count table into a clean
side-by-side year comparison, exactly the shape you'd want for a report.

---

## Show concluding results

```{r}
#| eval: false
survey_combined |>
  ggplot(aes(x = financial_wellness_cluster, y = income, fill = factor(year))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Income by Cluster: 2017 vs 2022",
       x = "Cluster", y = "Income ($)", fill = "Year") +
  theme_minimal()
```

This is the exact plot style from Day 1 — except now you've watched every
step of how the data underneath it got here.

---

## Back to your predictions

<div style="text-align: center; font-size: 1.3em; margin-top: 1.5em;">
Which of your guesses from the start of class came true?

<br><br>
What did we run into that nobody predicted?
</div>

---

## A few honest cautions

- Every cleaning choice was also an analytical choice — a different 3-level
  collapse, a different age-bucket rule, a different imputation method
  could all shift what we find downstream
- This dataset is synthetic — today's numbers are illustrative, not a claim
  about real Canadians
- Later this semester: missing-data methods and statistical tests that let
  you make these calls more rigorously than by eyeballing
