---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 7: Visualization Recap + Midterm Review"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "October 1, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))

# Scatterplot data with a deliberate outlier for the interpretation exercise
set.seed(42)
subgroup_data <- mpg |>
  filter(class %in% c("compact", "suv", "2seater")) |>
  bind_rows(tibble(
    manufacturer = "outlier", model = "outlier", displ = 2.0, hwy = 12,
    cyl = 4, trans = "auto", drv = "f", cty = 10, fl = "p",
    class = "compact", year = 2008
  ))
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))

set.seed(42)
subgroup_data <- mpg |>
  filter(class %in% c("compact", "suv", "2seater")) |>
  bind_rows(tibble(
    manufacturer = "outlier", model = "outlier", displ = 2.0, hwy = 12,
    cyl = 4, trans = "auto", drv = "f", cty = 10, fl = "p",
    class = "compact", year = 2008
  ))
```

## What we've covered

In the past two weeks, we have covered:

- **Lecture 4**: the grammar of graphics with data, aesthetics, geoms, facets, stats, positions, coordinates
- **Lecture 5**: exploratory data analysis with variation, covariation, outliers, missing data, confounding
- **Lecture 6**: design principles with visual cues, honest axes, showing the data, easing comparisons

::: {.callout-note title="Today's Agenda"}
We are going to review these concepts from above for next week's midterm.
:::

# Aesthetics and geoms

## Aesthetics at a glance {.smaller}

| Aesthetic | Controls | Mapping example |
|---|---|---|
| `x`, `y` | position | `aes(x = displ, y = hwy)` |
| `color` | outline / point color | `aes(color = class)` |
| `fill` | interior color (bars, areas) | `aes(fill = class)` |
| `shape` | point symbol (≤6 levels) | `aes(shape = drv)` |
| `size` | point/line size (continuous only) | `aes(size = cty)` |
| `alpha` | transparency (continuous only) | `aes(alpha = year)` |

::: {.callout-important}
Inside `aes()` = **mapping**, which is driven by data and you get a legend. 

Outside `aes()`= **setting**, which is essentially a fixed constant, no legend.
:::

## Mapping vs. setting {.smaller}

::: {.callout-tip title="Mapping"}
Ties an aesthetic to a variable in your data. The value changes per row,
and ggplot2 draws a legend. Written **inside** `aes()`.
:::

::: {.callout-note title="Setting"}
Gives an aesthetic one fixed value for every point, no matter what the data says. No legend, since there's nothing to decode. Written **outside** `aes()`.
:::

```r
# Mapping: color depends on class, varies per point, gets a legend
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

# Setting: every point is blue, no legend
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```

::: {.callout-caution title="Trap"}
`geom_point(aes(color = "blue"))` looks like a setting but is written inside `aes()`, so it's a mapping to the constant string `"blue"`, producing one default-palette color and a stray one-entry legend.
:::

## Try it yourself

::: {.callout-tip title="Run the code below"}
Predict what happens if you move `color = "blue"` outside `aes()`. 
:::

```{webr}
#| exercise: mapping_vs_setting

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = "blue"))
```


## Geoms at a glance {.smaller}

| Chart type | Geom | Notes |
|--|--|----|
| Scatterplot | `geom_point()` | default position `"identity"` |
| Jittered scatter | `geom_jitter()` | same as `geom_point(position = "jitter")` |
| Trend line | `geom_smooth()` | `se = FALSE` drops the confidence band |
| Histogram | `geom_histogram()` | needs `binwidth`; continuous x |
| Bar chart (counts) | `geom_bar()` | uses `stat_count()` automatically |
| Bar chart (values) | `geom_col()` | uses `stat_identity()`, needs `y` |
| Boxplot | `geom_boxplot()` | five-number summary + outliers |
| Frequency polygon | `geom_freqpoly()` | line version of a histogram |

# facets, positions, stats

## Facets {.smaller}

| Function | Splits by | Layout |
|---|---|---|
| `facet_wrap(~var)` | one variable | wraps into a grid automatically |
| `facet_grid(rows ~ cols)` | one or two variables | fixed rows/columns |
| `facet_grid(var ~ .)` | one variable | forced into rows |
| `facet_grid(. ~ var)` | one variable | forced into columns |

::: {.callout-note}
`scales = "free"` (or `"free_x"`/`"free_y"`) lets each panel use its own axis range — easier to read within a panel, but panels are no longer directly comparable.
:::

## Position adjustments {.smaller}

| Position | Effect | Best for |
|---|---|---|
| `"identity"` | draws values exactly where they fall | points; overlapping bars need `alpha` |
| `"stack"` | segments piled up | default for bars |
| `"fill"` | stacked, normalized to 100% | comparing **proportions** |
| `"dodge"` | segments side by side | comparing **raw counts** |
| `"jitter"` | small random noise added | reducing overplotting in scatterplots |



## Stats and `after_stat()` {.smaller}

Every geom has a default **stat** that computes what actually gets plotted. For `geom_bar()`, that default is `count`: the number of rows in each group. You never have to ask for it explicitly.

::: {.callout-note title="`after_stat()`"}
Lets you reach into the stat's output and ask for a *different* computed variable instead of the default:

| Variable | Definition |
|---|---|
| `count` | number of observations per group (default) |
| `prop` | proportion of the total each group represents |
| `density` | count standardized so the area under the curve sums to 1 |

:::

``` {r}
#| eval: false
ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) + geom_bar()
```

## Count vs. proportion vs. density

```{r}
#| fig-height: 3
#| layout-ncol: 3
#| fig-width: 3.2
ggplot(diamonds, aes(x = cut)) +
  geom_bar() +
  labs(title = "count (default)")

ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar() +
  labs(title = "prop")

ggplot(diamonds, aes(x = price, y = after_stat(density))) +
  geom_histogram(binwidth = 500) +
  labs(title = "density")
```

::: {.callout-note}
Same data, same bars, three different y-axes. `count` shows raw group sizes; `prop` rescales so every bar's height reflects its share of the whole (summing to 1 across bars); `density` is more commonly used with `histograms`/`freqpolys` and re-scales for area rather than height, here it looks nearly identical to `prop` since `geom_bar()` bins are all width 1.
:::



# labels and scales

## Labels

```r
labs(
  title = "States the finding, not the axes",
  subtitle = "Optional supporting detail",
  x = "Engine displacement (L)",
  y = "Highway mpg",
  color = "Vehicle class",
  caption = "Data source"
)
```

::: {.callout-tip}
A good title tells the reader what to conclude. "Scatterplot of displ vs.
hwy" wastes the most prominent text on the slide.
:::

## Scales

| Function | Purpose |
|---|---|
| `coord_cartesian(xlim = ..., ylim = ...)` | zooms the view; refits nothing |
| `scale_x_continuous(limits = ...)` /`scale_y_continuous(limits = ...)` | discards data outside the range before any stat runs |

::: {.callout-caution title="Zooming"}
`coord_cartesian()` and scale `limits` both "zoom," but only one of them actually refits a `geom_smooth()` curve on less data.
:::

# Variation vs. covariation

## The distinction {.smaller}

::: {.callout-note title="Variation"}
How much a single variable's values differ from one observation to the next.
:::

::: {.callout-tip title="Covariation"}
The tendency for two or more variables to move together. If two variables covary, knowing one improves your guess about the other.
:::

| Comparison type | Tool |
|---|---|
| One numeric variable | histogram, density, boxplot |
| Categorical × numeric | boxplot, faceted histogram, colored freqpoly |
| Categorical × categorical | `geom_count()`, `geom_tile()` (heat maps) |
| Numeric × numeric | scatterplot, binned boxplots |


# Reading plots: practice

## Interpretation 1: boxplot

```{r}
#| echo: false
#| fig-height: 3.6
ggplot(subgroup_data, aes(x = class, y = hwy)) +
  geom_boxplot() +
  labs(x = "Vehicle class", y = "Highway mpg")
```

::: {.callout-tip title="Questions"}
a) Which class shows the most variation in `hwy`?
b) Is there evidence of covariation between `class` and `hwy`?
c) Does any class show a labeled outlier? How does a boxplot decide that?
d) Rank the three classes by median, best to worst mpg.
:::

## Interpretation 2: scatterplot

```{r}
#| echo: false
#| fig-height: 3.6
ggplot(subgroup_data, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Engine displacement (L)", y = "Highway mpg", color = "Class")
```

::: {.callout-tip title="Questions"}
a) Which class shows the most spread, and in which direction (x, y, or both)?
b) Within one class, does `hwy` fall as `displ` rises?
c) Do the three classes occupy separate regions, or do they overlap?
:::

# Live practice

## Exercise 1: aesthetics and geoms

::: {.callout-tip}
Using `subgroup_data`, build a scatterplot of `hwy` vs. `displ`. Map colour to `class`, and make the points partially transparent.
:::

```{webr}
#| exercise: practice_aes
if (!exists("subgroup_data")) {
  subgroup_data <- mpg |> filter(class %in% c("compact", "suv", "midsize"))
}
ggplot(subgroup_data, aes(x = displ, y = hwy)) +
  geom_point()
```

## Exercise 2: facets and positions

::: {.callout-tip}
Build a bar chart of `class` counts, filled by `drv`, with bars **side by side** rather than stacked. Then facet the same plot by `drv`.
:::

```{webr}
#| exercise: practice_facet_position

ggplot(subgroup_data, aes(x = class, fill = drv)) +
  geom_bar()
```

## Exercise 3: labels and scales

::: {.callout-tip}
Take the plot from Exercise 1 and add a title that states a finding, proper
axis labels, and a ColorBrewer palette.
:::

```{webr}
#| exercise: practice_labels

ggplot(subgroup_data, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7)
```

## Final tips before your exam

- Review the slides and textbook (*R for Data Science*).

- Go over the pre-lab and lab assignment solutions. 
  - Make sure that you understand the material.

- Review your cheat sheet
  - please write your own: An authentic formula sheet will be far more useful than a sheet that you have never read before, written by your friend.

- I still have office hours from 2 to 4pm on next Monday, October 5!



- Practice reading a plot cold: variation, covariation, outliers, labels.

