---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 3: The Whole Game - Day 2"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 17, 2026"
format: 
  live-revealjs:
    resources:
      - data/raw_survey_2017.csv
      - data/raw_survey_2022.csv
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
---

## Recap: Day 1 of Whole Game

- We examined some EDA (Exploratory Data Analysis): with variables like income, benefits, delayed paycheque difficulty, etc.
- We ended with three financial wellness clusters: Stressed, Coping, Comfortable
- Today: we look at how we actually got there , ... including the messy parts of data analytics ...

---

## Before we look at the raw data...


::: {.callout-caution title="Thinking Moment"}
What do you think could go wrong with real, raw survey data?
:::


::: {.callout-warning title="Potential Problems"}
What kinds of problems might it have that yesterday's polished plots didn't show?
:::


---

```{r set up}
#| include: false
library(tidyverse)
library(ggplot2)

survey_2017_raw <- read_csv("data/raw_survey_2017.csv") |> distinct()
survey_2022_raw <- read_csv("data/raw_survey_2022.csv", skip = 1) |> distinct()

benefits_2017_long <- survey_2017_raw |>
  pivot_longer(cols = starts_with("benefit_"), names_to = "benefit_type", values_to = "response") |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_, .groups = "drop")
survey_2017_clean <- survey_2017_raw |> select(-starts_with("benefit_")) |> left_join(benefits_2017_long, by = "respondent_id")

benefits_2022_long <- survey_2022_raw |>
  pivot_longer(cols = starts_with("benefit_"), names_to = "benefit_type", values_to = "response") |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_, .groups = "drop")
survey_2022_clean <- survey_2022_raw |> select(-starts_with("benefit_")) |> left_join(benefits_2022_long, by = "respondent_id")

clean_income <- function(x) {
  x <- str_trim(x)
  is_k <- str_detect(x, "K$")
  numeric_part <- x |> str_remove_all("[$,K]") |> as.numeric()
  if_else(is_k, numeric_part * 1000, numeric_part)
}
survey_2017_clean <- survey_2017_clean |> mutate(income = clean_income(income))
survey_2022_clean <- survey_2022_clean |> mutate(income = clean_income(income))

standardize_province <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(
    x %in% c("ontario", "on") ~ "Ontario",
    x %in% c("alberta", "ab") ~ "Alberta",
    x %in% c("british columbia", "bc") ~ "British Columbia",
    x %in% c("quebec", "qc") ~ "Quebec",
    TRUE ~ str_to_title(x)
  )
}
survey_2017_clean <- survey_2017_clean |> mutate(province = standardize_province(province))
survey_2022_clean <- survey_2022_clean |> mutate(province = standardize_province(province))

standardize_yn <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(x %in% c("yes", "y") ~ "Yes", x %in% c("no", "n") ~ "No", TRUE ~ NA_character_)
}
survey_2017_clean <- survey_2017_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))
survey_2022_clean <- survey_2022_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))

survey_2017_clean <- survey_2017_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult", "Somewhat difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Somewhat manageable", "Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))
survey_2022_clean <- survey_2022_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))

# age bucket alignment
survey_2017_clean <- survey_2017_clean |>
  mutate(age_bucket = case_when(
    age <= 30 ~ "18-30",
    age <= 40 ~ "30-40",
    age <= 50 ~ "40-50",
    age <= 65 ~ "50-65",
    TRUE ~ "65+"
  )) |>
  select(-age)

survey_2022_clean <- survey_2022_clean |>
  rename(age_bucket = age)

# column name alignment
survey_2017_clean <- survey_2017_clean |>
  rename(pay_delay_response = paycheque_delay_difficulty)
survey_2022_clean <- survey_2022_clean |>
  rename(pay_delay_response = twoweek_pay_delay_impact)

# benefit category alignment
survey_2017_clean <- survey_2017_clean |>
  mutate(benefit_preference = if_else(
    benefit_preference == "benefit_education",
    "benefit_other",
    benefit_preference
  ))
survey_combined <- bind_rows(survey_2017_clean, survey_2022_clean)
```


## The reveal: here's the actual raw file

```{r}
#| eval: false
survey_2017_raw <- read_csv("data/raw_survey_2017.csv")

# Note the skip = 1 -- 2022's file has a stray metadata line above the
# real header (a leftover from the export on the new survey platform)
survey_2022_raw <- read_csv("data/raw_survey_2022.csv", skip = 1)

glimpse(survey_2017_raw)
glimpse(survey_2022_raw)
```


---

## Duplicate responses (rows)

A few respondents show up more than once (simulating a double-submitted survey).

```{r}
#| eval: false
survey_2017_raw |> count(respondent_id, sort = TRUE) |> filter(n > 1)

survey_2017_raw <- survey_2017_raw |> distinct()
survey_2022_raw <- survey_2022_raw |> distinct()
```

::: {.callout-tips}
The function `distinct()` removes exact duplicate rows. It is always worth checking for *before* doing anything else), since duplicates quietly inflate every count.
:::


---

## Tidy data: benefits are stuck in wide format

```{r echo=TRUE} 
#| eval: false
survey_2017_raw |>
  select(respondent_id, starts_with("benefit_")) |>
  head()
```

::: {.callout-tip}
One respondent has a single true preference, but right now it's scattered across six separate columns. Before we can do anything useful with it, we need to reshape it with `pivot_longer()`.
:::



---

## Tidy data: `pivot_longer()` 

```{r}
#| eval: false
benefits_2017_long <- survey_2017_raw |>
  pivot_longer(
    cols = starts_with("benefit_"),
    names_to = "benefit_type",
    values_to = "response"
  ) |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(
    benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_,
    .groups = "drop"
  )

survey_2017_clean <- survey_2017_raw |>
  select(-starts_with("benefit_")) |>
  left_join(benefits_2017_long, by = "respondent_id")
```


```{r echo = FALSE}
#| eval: false
benefits_2022_long <- survey_2022_raw |>
  pivot_longer(
    cols = starts_with("benefit_"),
    names_to = "benefit_type",
    values_to = "response"
  ) |>
  mutate(chosen = response %in% c("Yes", "Y", "yes")) |>
  group_by(respondent_id) |>
  summarize(
    benefit_preference = if (any(chosen)) benefit_type[chosen][1] else NA_character_,
    .groups = "drop"
  )

survey_2022_clean <- survey_2022_raw |>
  select(-starts_with("benefit_")) |>
  left_join(benefits_2022_long, by = "respondent_id")
```

---

## Strings & expressions


::: {.callout-warning title="The format for income response is a mess!"}

Sometimes people are allowed to type intheir response however they like. For example, if someone has an annual income of $85,000, there are several ways they might enter that in the survey:

- `"85000"`
- `"$85,000"`
- `"85,000"`
- `"85K"`

So how do we take all these different formats and turn them into something we can actually work with in our analysis?
:::


## Recall from last week's survey

For Question 7 of the survey, I asked everyone what was the best sandwich?

::: {.callout-note title="Most Popular Answer"}

**Grilled Cheese** was among the most popular answers. But this is what I got from last week's survey response:

- Grilled cheese
- Grilled Cheese
- Grilled Cheese!
- grilled cheese
- grilled cheese with Hot sauce
- Grilled cheese sandwich
- Ham Grilled Cheese

:::




## Back to the Income Example

```{r echo=FALSE}
#| eval: false
survey_2017_clean |> distinct(income) |> slice_sample(n = 6)
```

::: {.callout-important title="Solution"}
Before income can be used in any calculation, the income data needs to be transformed into a real number.
:::


```{r}
#| eval: false
clean_income <- function(x) {
  x <- str_trim(x)
  is_k <- str_detect(x, "K$")
  numeric_part <- x |>
    str_remove_all("[$,K]") |>
    as.numeric()
  if_else(is_k, numeric_part * 1000, numeric_part)
}

survey_2017_clean <- survey_2017_clean |>
  mutate(income = clean_income(income))

survey_2022_clean <- survey_2022_clean |>
  mutate(income = clean_income(income))
```

---

## Strings & expressions: provinces

```{r echo=FALSE}
#| eval: false
survey_2017_clean |> distinct(province)
```

Same province, several spellings: `"Ontario"`, `"ON"`, `"ontario "`.

```{r}
#| eval: false
standardize_province <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(
    x %in% c("ontario", "on") ~ "Ontario",
    x %in% c("alberta", "ab") ~ "Alberta",
    x %in% c("british columbia", "bc") ~ "British Columbia",
    x %in% c("quebec", "qc") ~ "Quebec",
    TRUE ~ str_to_title(x)
  )
}

survey_2017_clean <- survey_2017_clean |> mutate(province = standardize_province(province))
survey_2022_clean <- survey_2022_clean |> mutate(province = standardize_province(province))
```

---

## Strings & expressions: Yes/No

```{r echo=FALSE}
#| eval: false
survey_2017_clean |> distinct(debt_overwhelmed)
```

We might get three variations for the same two answers: `"Yes"/"No"`, `"Y"/"N"`, `"yes"/"no"`.

```{r}
#| eval: false
standardize_yn <- function(x) {
  x <- str_trim(x) |> str_to_lower()
  case_when(
    x %in% c("yes", "y") ~ "Yes",
    x %in% c("no", "n") ~ "No",
    TRUE ~ NA_character_
  )
}

survey_2017_clean <- survey_2017_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))
survey_2022_clean <- survey_2022_clean |> mutate(debt_overwhelmed = standardize_yn(debt_overwhelmed))
```

---

## Missing data


::: {.callout-warning title="Where did missing data come from?"}

Sometimes we don't always get all the complete data set that we want. 

- People tend to skip questions that they don't feel comfortable answer.
- perhaps people forgot to fill in one of the questions/pages of survey.
- maybe people gave some weird looking answer that is unusable (e.g $*@2#$1aE&)
- or sometimes these questions are simply irrelevant to them

:::

---



## Two options: 1) Delete the row

If a respondent didn't answer, we can just remove them from the analysis. :)

::: {.callout-tip title="This is cool, but..."}

While this is a simple method, you will lose the respondents. 

- if the people who skipped the question aren't random, the ones left behind don't represent everyone fairly.

:::


```{r}
#| eval: false
survey_2017_nomiss <- survey_2017_clean |>
  drop_na(income, savings_pct,
          debt_overwhelmed,
          emergency_fund_likelihood)
```

---

## Two options: 2) Imputation

Instead of throwing the whole response away, we can guess a reasonable value to put in its place.

- like using the average of everyone else's answer. (e.g. mean or median)

```{r}
#| eval: false
survey_2017_clean |> mutate(income = ifelse(is.na(income), median(income, na.rm = TRUE), income))
```

::: {.callout-tip title="This is cool, but..."}

Now some of your "data" is really just a guess. If you fill in too many values this way, your data can look more similar than it actually is.

:::


---

```{r echo=FALSE}
#| eval: false
survey_2017_clean |> distinct(emergency_fund_likelihood)
survey_2022_clean |> distinct(emergency_fund_likelihood)
```

## Mismatched response

::: {.callout-warning title="Too many response levels"}

Sometimes survey designers choose a set of response options one year, then decide to change them the next. For example, in our financial wellness data, the survey asked questions of *How difficult is it to obtain $2,000 of emergency fund?*:

- in 2017, there are 7 levels: *Very difficult*, *Difficult*, *Somewhat difficult*, *Neither*, *Somewhat manageable*, *Manageable*, *Very manageable* 
- in 2022, there are now only 5 levels: *Very difficult*, *Difficult*, *Neither*, *Manageable*, *Very manageable* 

So what do you do? Neither of them directly match each other. And this is difficult for a year-to-year comparative analysis
:::


## Potential Solution

::: {.callout-note title="Compromise"}

We can collapse both down to 3 levels — *Difficult*, *Neither*, *Manageable.*

:::

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult", "Somewhat difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Somewhat manageable", "Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))

survey_2022_clean <- survey_2022_clean |>
  mutate(emergency_fund_3lvl = case_when(
    emergency_fund_likelihood %in% c("Very difficult", "Difficult") ~ "Difficult",
    emergency_fund_likelihood == "Neither" ~ "Neither",
    emergency_fund_likelihood %in% c("Manageable", "Very manageable") ~ "Manageable",
    TRUE ~ NA_character_
  ))
```



## Before: 7 Levels vs 5 Levels

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
bind_rows(
  survey_2017_clean |> mutate(year = 2017),
  survey_2022_clean |> mutate(year = 2022)
) |>
  filter(!is.na(emergency_fund_likelihood)) |>
  mutate(emergency_fund_likelihood = factor(
    emergency_fund_likelihood,
    levels = c("Very difficult", "Difficult", "Somewhat difficult", "Neither",
               "Somewhat manageable", "Manageable", "Very manageable")
  )) |>
  ggplot(aes(x = emergency_fund_likelihood, fill = emergency_fund_likelihood)) +
  geom_bar(aes(y = after_stat(count / sum(count))), color = "white") +
  facet_wrap(~ year, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Emergency Fund Readiness: Raw Response Levels", x = NULL, y = "Percentage") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))
```

Different number of categories, different wording. It is not easy to tell how people have changed over time by comparing these two bar plots side by side as-is.


## After:

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
bind_rows(
  survey_2017_clean |> mutate(year = 2017),
  survey_2022_clean |> mutate(year = 2022)
) |>
  filter(!is.na(emergency_fund_3lvl)) |>
  mutate(emergency_fund_3lvl = factor(emergency_fund_3lvl,
                                       levels = c("Difficult", "Neither", "Manageable"))) |>
  ggplot(aes(x = emergency_fund_3lvl, fill = emergency_fund_3lvl)) +
  geom_bar(aes(y = after_stat(count / sum(count))), color = "white") +
  facet_wrap(~ year) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Emergency Fund Readiness: 2017 vs 2022", x = NULL, y = "Percentage") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

Now that both years use the same 3 levels, we can finally put them side by side and ask: did things get better, worse, or stay the same?

---

## Different ways of asking the same question

```{r echo=FALSE}
#| eval: false
names(survey_2017_clean)
names(survey_2022_clean)
```

::: {.callout-warning title="Mismatched question: same question, different name"}

Sometimes a question gets worded differently across years, even though it's really asking the same thing.

- in 2017, the survey asked *How difficult was it for you if your paycheque is delayed?*
  - stored as `paycheque_delay_difficulty`
- in 2022, the question became *If there is a two week delay in your paycheque arrival, how impactful would it be to your finances?*
  - stored as `twoweek_pay_delay_impact`

Same underlying question, different wording and different column name — `rename()` fixes this in one line each.

:::

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  rename(pay_delay_response = paycheque_delay_difficulty)

survey_2022_clean <- survey_2022_clean |>
  rename(pay_delay_response = twoweek_pay_delay_impact)
```

---

## Mismatched responses: Age

```{r}
#| eval: false
survey_2017_clean |> select(age) |> head()
survey_2022_clean |> select(age) |> head()
```

::: {.callout-warning title="Are you willing to share your age?"}

Sometimes the same underlying question gets *measured* differently across years, even when the intent is identical. 

- in 2017, respondents gave an exact age
  - stored as `age`, e.g. `34`
- in 2022, respondents picked from age brackets instead
  - stored as `age`, e.g. `"30-40"`

- Now we have the same underlying question, but a different level of precision
  — to compare over the years, everyone needs to speak the same bucket language.

:::

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(age_bucket = case_when(
    age <= 30 ~ "18-30",
    age <= 40 ~ "30-40",
    age <= 50 ~ "40-50",
    age <= 65 ~ "50-65",
    TRUE ~ "65+"
  ))

survey_2022_clean <- survey_2022_clean |>
  rename(age_bucket = age)
```


---

## Removed benefit option: aligning categories

```{r}
#| eval: false
survey_2017_clean |> distinct(benefit_preference)
survey_2022_clean |> distinct(benefit_preference)
```

"Educational funding" existed as a choice in 2017, but the 2022 platform
dropped it. To compare years fairly, we fold anything not offered in both
years into a shared **"Other"** category.

```{r}
#| eval: false
survey_2017_clean <- survey_2017_clean |>
  mutate(benefit_preference = if_else(
    benefit_preference == "benefit_education",
    "benefit_other",
    benefit_preference
  ))
```

---

## Combine into one data set

```{r}
#| eval: false
survey_combined <- bind_rows(survey_2017_clean, survey_2022_clean)

glimpse(survey_combined)
```

::: {.callout-note title="Only possible now"}

Now that we have:

 — column names match (`rename()`)
 - category sets match (the "Other" recoded)
 - response scales match (the 3-level of likert responses in paycheque question)
 
Every fix so far existed to make this one line possible.

:::

---

## Too much info in one column

```{r echo=FALSE}
#| eval: false
survey_combined |> select(respondent_id) |> head()
```

::: {.callout-warning title="One column doing two jobs"}

The variable `respondent_id` looks like a single value, but it's actually smuggling in two separate pieces of information at once.

- the year the response came from, e.g. `2017`

- the respondent's number within that year, e.g. `0001`

Both are jammed into one string: `"2017-0001"`. If we ever want to work with just the year, or just the respondent number, we're stuck first pulling them apart.

:::

```{r}
#| eval: false
survey_combined <- survey_combined |>
  separate(respondent_id, into = c("survey_year", "respondent_num"), sep = "-")

survey_combined |>
  select(survey_year, respondent_num, everything()) |>
  head()
```


```{r echo=FALSE}
#| eval: false
survey_combined |>
  count(financial_wellness_cluster, year) |>
  pivot_wider(names_from = year, values_from = n)
```

---

## Show concluding results

```{r echo=TRUE}
#| fig-width: 9
#| fig-height: 5
survey_combined |>
  ggplot(aes(x = financial_wellness_cluster, y = income, fill = factor(year))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Income by Cluster: 2017 vs 2022",
       x = "Cluster", y = "Income ($)", fill = "Year") +
  theme_minimal()
```

This is the exact plot style from Day 1, except now you've watched every step of how the data underneath it got here.

---

---

## Back to your predictions

::: {.callout-important title="Discussion"}

Which of your guesses from the start of class came true?

What did we run into that nobody predicted?

:::

::: {.callout-note title="Document your decisions"}

A results plot without its cleaning log is a claim without evidence. For any real report, this deserves its own short section (*ahem your final report*):

- Which rows/columns were dropped, and why
- Every recoding decision (e.g. collapsing levels and combining buckets, etc.)
- What's imputed vs. observed, clearly marked

:::

---

## A few honest cautions

::: {.callout-caution title="Keep in mind"}

- Every cleaning choice was also an analytical choice — a different 3-level collapse, a different age-bucket rule, a different imputation method could all shift what we find downstream
- This dataset is synthetic — today's numbers are illustrative, not a claim about real Canadians
- Later this semester: missing-data methods and statistical tests that let you make these calls more rigorously than by eyeballing

:::


::: {.callout-tip title="Next Week"}
Visualization!
:::
