---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 4: Visualization - Day 1"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 22, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
    - ggrepel
    - patchwork
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
```

## Where we are

In this week, I will give you a working template for plots. This chapter explains
*why* that template works. 

By the end you should be able to describe **any** ggplot2 plot in terms of seven components:  a dataset, a geom, a set of mappings, a stat, a position adjustment, a coordinate system, a faceting scheme and theme.

::: {.callout-note title="My Plan"}

I hope that once you've seen the seven slots, these plots will stop looking like new syntax to memorize.

:::


## Setup

```{r}
#| eval: false
library(tidyverse)
```

We'll work mostly with two datasets that ship with ggplot2:

- `mpg` — 234 car models, 11 variables
- `diamonds` — ~54,000 diamonds

```{r}
mpg |> select(manufacturer, model, displ, hwy, class, drv) |> head(4)
```


# Data set

## Data

Every `ggplot()` call starts with the same first argument: a data frame (or
tibble).

```{r}
#| eval: false
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy))
```

`mpg` fills in `<DATA>` from the template. Nothing gets drawn until ggplot2 knows *which* data frame to pull the variables `displ` and `hwy` out of the mappings in `aes()`.


## Data can change per layer

`ggplot(data = ...)` sets the **default** data for every layer. Any geom can override it with its own `data` argument:

```{r}
#| eval: false
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(data = mpg |> filter(class == "2seater"), color = "red")
```

The first layer inherits `mpg` from `ggplot()`; the second layer supplies its own filtered data frame instead. Same idea you'll see again in the Layers section:

- data is a slot you can fill once globally, or override locally.


## The pipe: `|>`

The symbol `|>` takes what's on its left and feeds it in as the **first argument** of the function on its right.

```{r}
#| eval: false
mpg |> select(manufacturer, model, displ, hwy)

# is exactly the same as:
select(mpg, manufacturer, model, displ, hwy)
```


## Analogy

::: {.callout-tip title="If you like math..."}

Think of a two-variable function $f(x, y)$.

The pipe works like plugging a value in for one of the arguments. Piping `5` into $y$ gives you $f(x, 5)$. Essentially you've fixed one input and left the rest open.

:::

::: {.callout-caution title="Connecting back to coding"}

When we write `mpg |> filter(...) |> select(...)`, we're doing the same thing: treating `mpg` like the input we're plugging in, and `filter(...)`, `select(...)` like functions waiting to receive it as their first argument.

:::


## The Pipe Chain

You can chain several steps and get something that reads top-to-bottom, roughly like "take `mpg`, *then* filter it, *then* select these columns", instead of nesting function calls inside each other.

```{r}
#| eval: false
mpg |>
  filter(class == "suv") |>
  select(model, displ, hwy) |>
  arrange(desc(hwy))
```

::: {.callout-note}
This is R's native pipe (base R ≥ 4.1). You may also see `%>%` in older code or tutorials. They serve the same purpose.
:::



# Aesthetic mappings

## Mapping a variable to an aesthetic

An **aesthetic** is a visual property of the marks on the plot: position, color, shape, size, transparency.

`aes()` says "let this variable *control* that property."

```{r}
#| fig-height: 3.4
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Not every aesthetic suits 

You can change the shape of scatterplots from circles to triangles, stars, squares, etc. 

But `shape` only supplies six distinct symbols before it gives up: the seventh category simply goes unplotted, with a warning.

```{r}
#| fig-height: 3
#| warning: true
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```


## Problems with shape levels

::: {.callout-warning}
`class` has seven levels but `shape` can only display six. SUVs (62 rows) silently disappear from the plot above. You should always read the warnings `ggplot2` gives you, they're telling you data went missing from the picture.
:::

```{r}
#| fig-height: 3
#| warning: true
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_point()
```


## Point customization

::: {.callout-warning title="Checkpoint"}

What do you think `size` and `alpha` does?

:::

- `size` changes the size of the dot (smaller-larger)

- `alpha` changes the transparency of the dot (lighter-darker)

- These two aesthetics are *ranked*.

## Ordered aesthetics


```{r}
#| fig-height: 3
#| warning: true
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()
```

::: {.callout-caution title="Mapping an unordered category"}
`class` (e.g. 2seater, compact, midsize, ...) has no inherent order. Mapping it to `size` or `alpha` visually implies "bigger/darker = more," which is misleading for a purely categorical variable. Save these aesthetics for variables that are genuinely ordered or continuous.
:::

## Mapping vs. setting

Inside `aes()` = mapping, driven by data, gets a legend.
Outside `aes()` = setting, a fixed constant, no legend.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue", size = 2, shape = 17)
```

## Setting values

When you set an aesthetic you supply a raw value:

| Aesthetic | Value |
|---|---|
| `color` | colour name `"blue"` or hex string `"#E81B23"` |
| `size`  | point size in mm, `size = 1` |
| `shape` | an integer 0–25, `shape = 1` |
| `alpha` | opacity from 0 to 1 |

Shapes 0–14 are hollow (border from `color`), 15–20 are solid (filled with
`color`), and 21–25 take both a `color` border and a `fill`.

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Draw `hwy` vs. `displ` with pink filled triangles.
2. Why does the code below *not* produce blue points?
3. What does `stroke` do, and which shapes respond to it? (`?geom_point`)
4. What happens with `aes(color = displ < 5)`?
:::

```{r}
#| eval: false
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```

```{webr}
#| exercise: aes_mappings

```

# Geometric objects

## A geom is how the data gets drawn

Same data, same mappings, different geom:

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4.5
#| fig-height: 3.5
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4.5
#| fig-height: 3.5
ggplot(mpg, aes(displ, hwy)) +
  geom_smooth()
```
:::
::::

## Geoms accept different aesthetics

A line has no shape; a point has no linetype. `ggplot2` quietly ignores aesthetics a geom can't use.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) +
  geom_smooth()
```

Mapping a discrete variable splits the geom into one object per level.

## Grouping without a legend

`group` splits the data the same way, but adds no legend and no visual
distinction.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv))
```

::: {.callout-note}
`group` is what mapping a discrete variable does *implicitly* behind the
scenes. You should use it when you want the splitting behaviour without the legend or the visual distinction that `color` or `linetype` would add.
:::

## Layers stack

Multiple geoms in one plot. Mappings in `ggplot()` are global; mappings in a
geom are local to that layer.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE)
```

## Layers can have their own data

Useful for highlighting a subset:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "grey60") +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red", size = 2
  )
```

## Different geoms

```{r}
#| fig-height: 3
#| layout-ncol: 3
#| fig-width: 3.2
ggplot(mpg, aes(x = hwy)) + geom_histogram(binwidth = 2)
ggplot(mpg, aes(x = hwy)) + geom_density()
ggplot(mpg, aes(x = hwy)) + geom_boxplot()
```

The histogram and density show *bimodality* and *right skew*; the boxplot compresses that into a summary but flags outliers.

## Common geoms at a glance {.smaller}

| Chart type | Geom | What it draws |
|---|---|---|
| Line chart | `geom_line()` | Connects points in order of the x-value — good for trends over a continuous sequence (e.g. time) |
| Bar plot | `geom_bar()` | Counts observations per category and draws bar height as the count (uses `stat_count()`) |
| Bar plot (pre-counted) | `geom_col()` | Bar height comes directly from a `y` column you already have — no counting |
| Histogram | `geom_histogram()` | Bins a *continuous* variable and draws bar height as the count per bin |
| Boxplot | `geom_boxplot()` | Five-number summary (min, Q1, median, Q3, max) plus outliers, per group |
| Area chart | `geom_area()` | Like `geom_line()`, but fills the region under the line |



## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Which geom draws a line chart? A boxplot? A histogram? An area chart?
2. What does `show.legend = FALSE` do, and why might you want it?
3. What does `se` control in `geom_smooth()`?
4. Recreate a plot with points colored by `drv` and *one* smooth curve fitted
   to the whole dataset (not one per group).
:::

```{webr}
#| exercise: geoms

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(aes(group = drv))
```

# Facets

## `facet_wrap()`: one variable

What if we are interested in how each class of variables behave? We can split the plot into panels with one per level.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```

## `facet_grid()`: two variables

A two-sided formula, `rows ~ cols`.

```{r}
#| fig-height: 3.4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

Empty panels mean that combination doesn't occur in the data.

## Free scales

By default every panel shares one scale, which is what makes panels comparable. Release that with `scales`:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl, scales = "free")
```


## Free scales

::: {.callout-note title="Customization"}
`"free_x"`, `"free_y"`, or `"free"` let each row/column pick its own range. 

- This is easier to read within a panel, but panels are no longer directly comparable
to each other. 

:::

::: {.callout-warning title="Checkpoint"}

If I want to share a plot with someone that has no idea about the data set, should I use free scales or shared scales?

:::


## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What happens if you facet on a continuous variable like `cty`?
2. What does `.` mean in `facet_grid(drv ~ .)`?
:::

```{webr}
#| exercise: facets

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~___)
```


# Statistical transformations

## Where does `count` come from?

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

`count` isn't a column in `diamonds`. The `geom_bar` computed it.

## Stats

A **stat** is the algorithm a geom runs on your data before drawing.

- bar charts, histograms, freqpolys → bin and count
- smoothers → fit a model, plot predictions
- boxplots → compute a five-number summary

Every geom has a default stat; every stat has a default geom. `?geom_bar`
tells you it uses `stat_count()`, and lists the variables it computes.

## Overriding the stat

When you already have the counts, switch to the identity stat (or use
`geom_col()`):

```{r}
#| fig-height: 3
diamonds |>
  count(cut) |>
  ggplot(aes(x = cut, y = n)) +
  geom_bar(stat = "identity")
```

## Using a computed variable

`after_stat()` reaches into the variables the stat produced. The `prop` means proportion.

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```



## `group` issue

::: {.callout-important}
Without `group = 1`, each bar is treated as its own group and every proportion comes out as 1, then there's nothing to divide by. Try deleting `group = 1` in the live cell below and see what happens.
:::

```{webr}
#| exercise: after_stat

ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```




## Calling a stat directly

Sometimes the summary *is* the point, and naming the stat says so:

```{r}
#| fig-height: 3
ggplot(diamonds) +
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min, fun.max = max, fun = median
  )
```


## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. If you swap `geom_bar()` for `geom_col()` in the plot below without changing anything else, what error (if any) do you think you'll get? Try it and see if you were right.
2. Which diamond `cut` do you think has the *smallest* proportion in the data set?
:::

```{webr}
#| exercise: stats

ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```

# Position adjustments

## `color` vs. `fill` for bars

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = drv, color = drv)) + geom_bar()
ggplot(mpg, aes(x = drv, fill = drv)) + geom_bar()
```

## Stacking is a position adjustment

Map `fill` to a second variable and the bars stack automatically:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar()
```

## The four options

::: {.callout-note title="Stack"}
**`"stack"`** — the default for bars. Segments pile on top of each other in full-height, full-count form.
:::

::: {.callout-tip title="Fill"}
**`"fill"`** — stack, then normalize each bar to the same height. Best for comparing **proportions** across groups, since every bar reaches 100%.
:::

::: {.callout-important title="Dodge"}
**`"dodge"`** — segments sit side by side instead of stacking. Best for comparing **individual counts** directly, since bar heights stay on the original count scale.
:::

::: {.callout-caution title="Identity"}
**`"identity"`** — draws each value exactly where it falls, with no adjustment. For bars this usually means overlapping rectangles (rarely what you want); for points, it's the default position since points have no width to adjust.
:::

## Exercise

::: {.callout-tip}
Change `position = "stack"` below to `"fill"`, `"dodge"`, and `"identity"`. Watch how the same data tells a different story each time.
:::

```{webr}
#| exercise: position_types

ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "stack")
```


## Overplotting

`hwy` and `displ` are rounded, so points land on a grid and pile up. This
scatterplot shows 126 visible points for 234 observations. The points are **overlapping**!

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Jittering

`position = "jitter"` adds a little random noise. Less accurate at small
scales, far more honest at large ones.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter()
```

`geom_jitter()` is shorthand for `geom_point(position = "jitter")`.


## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What's wrong with a plain `geom_point()` of `cty` vs. `hwy`? Fix it below.
2. Is there any difference between the default and `position = "identity"`
   for points?
3. Try `geom_jitter(width = 0.2, height = 0.5)` to control the amount of jitter.
4. Compare `geom_jitter()` and `geom_count()`.
:::

```{webr}
#| exercise: positions

ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
```

# Coordinate systems

## The default is Cartesian

x and y act independently. Two alternatives come up often enough to know:

- `coord_quickmap()`: fixes the aspect ratio for geographic data
- `coord_polar()`: bends the Cartesian plane into polar coordinates

```{r}
#| eval: false
nz <- map_data("nz")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```

## Polar coordinates

A bar chart and a Coxcomb chart are the same plot in different coordinates:

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4
bar <- ggplot(diamonds) +
  geom_bar(aes(x = clarity, fill = clarity), show.legend = FALSE, width = 1) +
  theme(aspect.ratio = 1)

bar + coord_flip()
bar + coord_polar()
```

# Putting it together

## The extended template

```r
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

There are seven slots. In practice you only fill in three: `data`, `mappings`, `geom`, because `ggplot2` supplies sensible defaults for the rest.


## Summary

- **Aesthetics** map variables to visual properties; setting differs from mapping
- **Geoms** decide how marks are drawn; layers stack and can override data and mappings
- **Facets** split one plot into comparable panels
- **Stats** compute the values a geom actually plots
- **Positions** resolve overlap: stack, fill, dodge, jitter
- **Coordinates** redefine what x and y mean

Still to come: **themes**, in Chapter 11 (next week).

