---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 10: Exploratory Data Analysis"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "Date, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
    - ggrepel
    - patchwork
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
smaller <- diamonds |> filter(carat < 3)
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
smaller <- diamonds |> filter(carat < 3)
```

## The EDA loop

EDA is not a procedure with a checklist. It's a cycle you run until the
questions stop being interesting:

1. **Generate questions** about your data
2. **Search for answers** by visualizing, transforming, and modelling
3. **Refine** your questions, or generate new ones from what you found

Most of what you produce during EDA gets thrown away. That's fine — the
output is understanding, not plots.

::: {.notes}
Emphasize that dead ends are expected and not failure. Students often think
EDA should look tidy in retrospect.
:::

## Why it's not optional

Even when the research question arrives fully formed, you still have to
know whether the data can answer it.

Data cleaning is a *special case* of EDA: you're asking whether the data
meets your expectations, and using the same tools — visualization,
transformation, modelling — to find out.

## Setup

```{r}
#| eval: false
library(tidyverse)
```

Two guiding questions carry most of the weight:

1. What **variation** occurs *within* my variables?
2. What **covariation** occurs *between* my variables?

::: {.callout-note}
`smaller` (`diamonds` filtered to `carat < 3`) is already available in every
live cell below, along with `mpg` and `diamonds`.
:::

# Variation

## Variation

**Variation** is the tendency of a variable's values to change from one
measurement to the next.

Measure anything continuous twice and you get two answers — even for
quantities that are genuinely constant. Add real differences between
subjects and across time, and every variable gets its own characteristic
pattern.

The way to see that pattern is to plot the distribution.

## A first look

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

## Binwidth changes what you can see

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Same data. A much narrower bin reveals structure the first plot hid entirely.

## Typical values: what to ask

Tall bars are common values, short bars rare ones, gaps are values never
observed. Turn that into questions:

- Which values are most common? Why?
- Which are rare? Does that match what you expected?
- Are there unexpected patterns? What could explain them?

For the carat histogram: why the spikes at whole carats and common
fractions? And why more diamonds just *above* each spike than just below?

::: {.notes}
The answer involves both cutting practice and pricing incentives — a good
illustration that some EDA questions need domain knowledge, not more code.
:::

## Clusters mean subgroups

When a distribution separates into clumps, ask:

- How are observations *within* a cluster alike?
- How do clusters differ from each other?
- What explains the split?
- Could the clusters be an artifact rather than real structure?

Many of these push you toward a relationship *between* variables.

## Unusual values hide in plain sight

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5)
```

The only hint of anything odd is how wide the x-axis is.

## Zoom in on the y-axis

```{r}
#| fig-height: 3.4
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

Now three unusual regions show up: near 0, around 30, and around 60.

::: {.callout-note}
**`coord_cartesian()` zooms** — the bars are still computed from *all* the
data, so you can safely zoom the y-axis to see rare bins without changing
what got counted. `ylim()`/`xlim()` instead **discard** any data outside the
limits before the stat runs, which can silently change your bar heights.
When you just want a closer look, reach for `coord_cartesian()`.
:::

## Pull the outliers out

```{r}
unusual <- diamonds |>
  filter(y < 3 | y > 20) |>
  select(price, x, y, z) |>
  arrange(y)

unusual
```

## What we found

`y` is a dimension in millimeters. A diamond cannot be 0 mm wide, so those
rows are **missing data encoded as zero** — something no search for `NA`
would ever have surfaced.

The 32 mm and 59 mm entries are implausible too: over an inch across, but
priced like ordinary stones.

::: {.callout-important}
This is the core payoff of EDA: visualizing the data surfaced a data-quality
problem that a search for `NA` never would have. Always sanity-check units
and physically-plausible ranges, not just missingness flags.
:::

::: {.callout-caution}
Run your analysis with and without the outliers. If it barely moves and you
can't explain them, dropping them is defensible. If it moves a lot, find the
cause and **disclose the removal** in your write-up — don't drop values
without justification just because they're inconvenient.
:::

## Two ways to handle them

```{r}
#| eval: false
# Option 1: drop the whole row — usually too blunt
diamonds2 <- diamonds |> filter(between(y, 3, 20))

# Option 2: replace the bad value with NA — usually better
diamonds2 <- diamonds |>
  mutate(y = if_else(y < 3 | y > 20, NA, y))
```

One invalid measurement doesn't invalidate the rest of the observation. Apply
option 1 across every variable in messy data and you may have nothing left.

## Missing values in plots

ggplot2 won't guess where to put an `NA`, so it drops them and warns you.
Silence the warning with `na.rm = TRUE` — once you've understood why they're
missing.

```{r}
#| fig-height: 2.9
#| warning: true
diamonds2 <- diamonds |> mutate(y = if_else(y < 3 | y > 20, NA, y))

ggplot(diamonds2, aes(x = x, y = y)) +
  geom_point()
```

## Missingness can be the signal

In `nycflights13::flights`, a missing `dep_time` means the flight was
cancelled. So compare scheduled departure times for cancelled and
non-cancelled flights:

```{r}
#| fig-height: 2.8
nycflights13::flights |>
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min  = sched_dep_time %% 100,
    sched_dep_time = sched_hour + (sched_min / 60)
  ) |>
  ggplot(aes(x = sched_dep_time)) +
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4)
```

::: {.callout-tip}
Missing values aren't always noise to drop. Sometimes *why* a value is
missing is itself the most informative variable in the dataset —
`is.na(dep_time)` here is doing more analytical work than `dep_time` itself.
:::

## The comparison isn't fair yet

Cancelled flights are vastly outnumbered, so the two curves aren't
comparable on a count scale.

That's a covariation problem — and the next section is about fixing exactly
this kind of comparison.

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Explore the distributions of `x`, `y`, and `z` in `diamonds`. Which is
   length, which width, which depth?
2. Explore `price`. Try a wide range of binwidths. Anything surprising?
3. How many diamonds are 0.99 carat vs. exactly 1 carat? Why the gap?
4. Compare `coord_cartesian()` with `xlim()`/`ylim()` when zooming into a
   histogram.
:::

```{webr}
#| exercise: variation

ggplot(diamonds, aes(x = x)) +
  geom_histogram(binwidth = 0.25)
```

# Covariation

## Covariation

If variation describes behavior *within* a variable, **covariation**
describes behavior *between* variables: the tendency for two or more
variables to move together.

The strategy depends on the variable types:

- categorical × numerical
- categorical × categorical
- numerical × numerical

## Categorical × numerical: the count problem

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = price)) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

Group sizes differ so much that shape differences are invisible.

## Switch the y-axis to density

**Density** is the count standardized so the area under each curve is 1.
Since it isn't a column in the data, `after_stat()` fetches it from the stat.

```{r}
#| fig-height: 3
ggplot(diamonds, aes(x = price, y = after_stat(density))) +
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

## A surprise

Fair diamonds — the *lowest* quality cut — appear to have the highest
average price.

Frequency polygons are busy, though. A boxplot makes the comparison cleaner:

```{r}
#| fig-height: 2.8
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```

::: {.callout-important}
This is a genuine paradox, not a plotting artifact — worse-cut diamonds
really do look more expensive here. Hold onto this; the "Patterns and
models" section resolves it once `carat` enters the picture.
:::

## Reordering unordered categories

`cut` is an ordered factor, so its order is meaningful. Most categorical
variables aren't.

::: {.callout-tip}
`fct_reorder(variable, by, fun)` sorts factor levels by a summary statistic
of another variable — almost always makes a categorical-vs-numeric plot
easier to read at a glance.
:::

```{r}
#| fig-height: 2.7
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = class, y = hwy)) + geom_boxplot()
ggplot(mpg, aes(x = fct_reorder(class, hwy, median), y = hwy)) + geom_boxplot()
```

## Flip when labels are long

Swap the x and y mappings — no extra function needed:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = hwy, y = fct_reorder(class, hwy, median))) +
  geom_boxplot()
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Improve the cancelled-vs-non-cancelled departure time plot.
2. Which variable best predicts diamond price? How is it related to `cut`?
   Use that to explain why poorer cuts look more expensive.
3. Compare swapping x/y with adding `coord_flip()`.
4. Try `geom_violin()`, a faceted `geom_histogram()`, a colored
   `geom_freqpoly()`, and a colored `geom_density()` on the same comparison.
   What does each do well and badly?
:::

```{webr}
#| exercise: cat_num

ggplot(diamonds, aes(x = cut, y = price)) +
  geom_violin()
```

## Categorical × categorical: counts

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()
```

Circle area shows how many observations fall at each combination.

## Or compute the counts, then tile

```{r}
#| fig-height: 3.2
diamonds |>
  count(color, cut) |>
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
```

::: {.callout-note}
For unordered categories, the **seriation** package can reorder rows and
columns to expose structure; **heatmaply** does interactive versions of the
same idea for larger tables.
:::

## Numerical × numerical

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point()
```

A positive, roughly exponential relationship — and a lot of overplotting.

## Transparency helps

```{r}
#| fig-height: 3.4
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point(alpha = 1 / 100)
```

Clusters at 1, 1.5, and 2 carats become visible.

## Bin in two dimensions

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(smaller, aes(x = carat, y = price)) + geom_bin2d()
ggplot(smaller, aes(x = carat, y = price)) + geom_hex()
```

`geom_bin2d()` uses rectangles, `geom_hex()` hexagons (needs the **hexbin**
package). Both map bin counts to fill.

## Or bin one variable and reuse boxplots

```{r}
#| fig-height: 3.2
ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

`cut_width(x, width)` slices `x` into equal-width bins. `cut_number()` makes
bins with roughly equal *counts* instead.

::: {.notes}
Boxplots look about the same regardless of how many points they summarize.
`varwidth = TRUE` makes box width proportional to group size, which is worth
mentioning here.
:::

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Frequency polygon instead of boxplot for the conditional distribution —
   what changes? When does `cut_width()` beat `cut_number()`?
2. Visualize `carat` partitioned by `price`.
3. How do price distributions compare for very large vs. very small diamonds?
4. Combine two techniques to show cut, carat, and price together.
5. Why is a scatterplot better than a binned plot for spotting 2-D outliers?
:::

```{webr}
#| exercise: num_num

ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_number(carat, 20)))
```

# Patterns and models

## Interrogating a pattern

A systematic relationship shows up as a pattern. When you spot one, ask:

- Could this be coincidence?
- How would you describe the relationship?
- How strong is it?
- What other variables might be involved?
- Does it hold within subgroups?

## Variation vs. covariation

Variation creates uncertainty. Covariation *reduces* it.

If two variables covary, values of one improve your predictions about the
other. If the covariation is causal — a special case — you can use one to
control the other.

## Models extract patterns

The cut/price paradox exists because carat is entangled with both. Fit a
model for price from carat, then look at what's left over.

```{r}
#| eval: false
library(tidymodels)

diamonds <- diamonds |>
  mutate(log_price = log(price), log_carat = log(carat))

diamonds_fit <- linear_reg() |>
  fit(log_price ~ log_carat, data = diamonds)

diamonds_aug <- augment(diamonds_fit, new_data = diamonds) |>
  mutate(.resid = exp(.resid))
```

Log-transform first, fit, then exponentiate the residuals back to a price-like
scale.

::: {.callout-note}
`tidymodels` is heavy and not part of the live-code package set for this
deck, so this example stays static — run it locally in RStudio if you'd
like to reproduce it yourself.
:::

## The paradox dissolves

```{r}
#| eval: false
ggplot(diamonds_aug, aes(x = cut, y = .resid)) +
  geom_boxplot()
```

With the carat effect removed, the expected relationship appears: **relative
to their size, better-cut diamonds cost more.**

Modelling gets its own treatment once you have wrangling and programming in
hand — but the idea is worth meeting here.

## Summary

- EDA is a **loop**: question → visualize/transform/model → better questions
- **Variation** within a variable: typical values, clusters, unusual values
- Outliers are sometimes errors, sometimes discoveries. Investigate before
  deciding; disclose what you removed
- **Covariation** between variables: pick the technique by variable type
- Overplotting has several fixes: alpha, 2-D binning, binning one axis
- **Models** let you strip out a dominant relationship to see what remains

Next: turning exploratory graphics into ones other people can read.
