---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 2: The Whole Game - Day 1"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 15, 2026"
format: 
  live-revealjs:
    resources:
      - data/survey_2017.csv
      - data/survey_2022.csv
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
execute:
  warning: false   # Hides warnings and package loading messages
  message: false
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}


## The Whole Game
Imagine learning baseball by drilling practices into pitching, hitting, and fielding for three months, before ever seeing a new game.

::: {.callout-warning title=""}
This sounds exhausting. Nobody learns baseball that way.
:::

::: {.callout-tip appearance="simple"}
Instead, you would play a [simplified whole game]{style="color:red;"} first.
- It is messy, imperfect, but real.
- You'd get a feel for what the sport is before drilling into any of the skills.
:::



## This course works the same way! :)
We are going to take a quick glimpse on what you are about to do in this course, over this week.


::: {.column width="48%"}

::: {.callout-note title="Today"}
Play the whole game once:

* An open question $\rightarrow$ some cool results
:::

:::

::: {.column width="48%"}
::: {.callout-tip title="Rest of semester"}
Drill each piece:

* Cleaning, transformations, modelling, communication
:::

:::

By the end of the course, you'll play the whole game again (*wink wink* this time completing the final project by yourself.



---
### {#no-header-slide data-menu-title="Hidden Title for Menu"}
A very common question that a client might ask is:

- We are going to give you a set of data, can you tell me something about it?

---

## Why does this matter?

- Financial stress affects health, relationships, and decision-making
- Understanding ***what predicts*** financial comfort helps policymakers, employers, and individuals

. . .

::: {.callout-note title="Today"}
A look at real survey data on Canadians' financial wellness.
:::


## The Data
::: {.callout-note title="Data Overview"}
Survey responses from Canadians regarding income, employment, financial habits, and workplace benefits.

* **Sample:** Representative cross-section of Canadians across all provinces
* **Scale:** 1000 total respondents across two survey waves (in 2017 & 2022)
* **Scope:** Captures demographic variables alongside self-reported financial stress metrics (survey responses)
:::

. . .

::: {.footnote}
*Note: This data set originates from a real research project conducted in partnership with external industrial partners. To preserve confidentiality, the data has been anonymized and structured for teaching purposes.*
:::


## 8 Survey Questions

1. What is your age?
2. Which province are you currently living in?
3. What is your annual income in dollars?
4. What percentage of your paycheck do you put into savings?
5. Are you currently overwhelmed by debt?
6. How manageable are you able to come up with $2,000 of emergency fund right now?
7. How manageable are you able to live if your paycheque arrives two weeks late?
8. What is the most important benefit that you seek from your job?


<!-- R Set-up -->
```{r setup}
#| include: false
library(tidyverse)
library(scales)

# Load CSV files (Ensure these are in your project/data folder)
survey_2017 <- read_csv("data/survey_2017.csv")
survey_2022 <- read_csv("data/survey_2022.csv")
survey <- bind_rows(survey_2017, survey_2022)

# Set Likert Levels
likert_levels <- c(
  "Very difficult", "Difficult", "Somewhat difficult", "Neither",
  "Somewhat manageable", "Manageable", "Very manageable"
)

# Mutate factors and buckets
survey <- survey %>%
  mutate(
    paycheque_delay_difficulty = factor(paycheque_delay_difficulty, levels = likert_levels, ordered = TRUE),
    emergency_fund_likelihood  = factor(emergency_fund_likelihood, levels = likert_levels, ordered = TRUE),
    financial_wellness_cluster = factor(financial_wellness_cluster, levels = c("Stressed", "Coping", "Comfortable")),
    paycheque_delay_score      = as.numeric(paycheque_delay_difficulty),
    emergency_fund_score       = as.numeric(emergency_fund_likelihood),
    age_bucket = cut(age, breaks = c(18, 25, 35, 45, 55, 66), labels = c("18-24", "25-34", "35-44", "45-54", "55-65"), right = FALSE, include.lowest = TRUE),
    income_bucket = cut(income, breaks = c(30000, 50000, 75000, 100000, 150000, Inf), labels = c("$30k-49k", "$50k-74k", "$75k-99k", "$100k-149k","$150k+"), right = FALSE, include.lowest = TRUE),
    province_bucket = case_when(
      province == "Alberta" ~ "AB",
      province == "British Columbia" ~ "BC",
      province == "Ontario" ~ "ON",
      province == "Quebec" ~ "QC",
      TRUE ~ "Other"
    )
  )

survey_2017 <- survey %>% filter(year == 2017)
survey_2022 <- survey %>% filter(year == 2022)
```


## Glimpse of Data
```{r}
head(survey_2017)
```

# Section 1: Demographics

---


## What does the sample look like?

::: {.callout-note title="What do you notice?"}
I am going to show you the demographic-response (i.e. age, income, and province) distributions from the 2017 survey results.  Try and answer the following questions

- Can you tell me if there are any obvious patterns? 
- Do you like the plots that I created? 
- Based on those 3 plots, can you tell me about the financial wellness of these groups of respondents?
:::



## Income Distribution

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = income)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 10000, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Income", x = "Income ($)", y = "Frequency") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 14)
```

---

## Age Distribution

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = age)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 5, fill = "darkorange", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 14)
```

---

## Province Distribution

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = province_bucket, fill = province_bucket)) +
  geom_bar(aes(y = after_stat(count / sum(count))), color = "white") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2") + # Beautiful predefined color palette
  labs(
    title = "Distribution of Respondents by Province",
    x = "Province",
    y = "Percentage",
    fill = "Province"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") # Hides redundant legend since x-axis is already labeled
```

---



## Checkpoint
For the rest of the lecture, try and think of these questions:

- Can you tell me if there are any obvious patterns? 

- Do you like the plots that I created? 

- Based on those 3 plots, can you tell me about the financial wellness of these groups of respondents?

## Age vs Income

This is an example of a *Linear Model*.
```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey_2017, aes(x = age, y = income)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Income vs. Age (2017)",
    x = "Age",
    y = "Income ($)"
  ) +
  theme_minimal(base_size = 14)
```


## Tell me about the patterns again?
```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey_2017, aes(x = age, y = income)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") + 
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title = "Income vs. Age (2017)",
    x = "Age",
    y = "Income ($)"
  ) +
  theme_minimal(base_size = 14)
```




# Section 2: Financial Wellness Clusters

---

## How We Found These Groups/Clusters

- We didn't set the rules manually: The patterns emerged directly from the survey data.
- *Unsupervised Learning:* We used algorithms to group respondents based on mathematical similarity across many variables at once.
  - Methods like *k-Nearest Neighbors (k-NN)* and *t-SNE* helped us reduce high-dimensional data into distinct clusters.
  - Don't worry! The math behind clustering is **out of scope** of this course, just think of it as a tool that finds hidden patterns in complex data. If you are interested, you should take *ST362* and *ST494* in future!


## Clustering Results
- Analyzed 100+ survey questions across 10+ years of longitudinal project data.
  - This is our *actual* work from the lab. I am only replicating a small subset of the data for this class.

- 3 groups proved to be the mathematical sweet spot (2 lost key details; 4 added unnecessary noise).
  1. Stressed
  2. Coping
  3. Comfortable

- Connects the data outputs directly to the client's goal of understanding Canadian financial stress and security.




# Section 3: Financial and Behavioural Responses

---

## Savings Rate by Cluster (2017 vs 2022)

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = financial_wellness_cluster, y = savings_pct, fill = factor(year))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Savings Rate by Cluster: 2017 vs 2022", x = "Cluster", y = "Savings (% of paycheque)", fill = "Year") +
  theme_minimal(base_size = 14)
```

---

## Paycheque Delay Difficulty

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = financial_wellness_cluster, fill = paycheque_delay_difficulty)) +
  geom_bar(position = "fill") +
  facet_wrap(~ year) +
  scale_y_continuous(labels = scales::percent) +
  # Generates a sequential dark-to-light blue shade across Likert levels
  scale_fill_brewer(palette = "Blues", direction = -1) + 
  labs(
    title = "How Each Group of People Feel About A Delayed Paycheque",
    x = "Financial Wellness Cluster",
    y = "Proportion",
    fill = "Likert Score"
  ) +
  theme_minimal()
```



---

## Most Important Benefit

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey %>% filter(!is.na(benefits_preference)), 
       aes(x = financial_wellness_cluster, fill = benefits_preference)) +
  geom_bar(position = "fill") +
  geom_text(
    aes(label = scales::percent(after_stat(prop), accuracy = 1), group = benefits_preference),
    stat = "count",
    position = position_fill(vjust = 0.5),
    color = "white",
    size = 3,
    fontface = "bold"
  ) +
  facet_wrap(~ year) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Workplace Benefits Preference by Cluster",
    x = "Financial Wellness Cluster",
    y = "Proportion",
    fill = "Benefit Preference"
  ) +
  theme_minimal(base_size = 14)
```



---

## Emergency Fund

```{r echo=FALSE}
#| fig-width: 9
#| fig-height: 5
ggplot(survey, aes(x = financial_wellness_cluster, y = paycheque_delay_score, fill = factor(year))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Paycheque Delay Difficulty by Cluster: 2017 vs 2022",
       subtitle = "Score: 1 = Very difficult ... 7 = Very manageable",
       x = "Cluster", y = "Difficulty Score", fill = "Year") +
  theme_minimal(base_size = 14)
```


---




## Exercise
Try plotting yourself!

```{webr}
library(ggplot2)
library(tidyverse)
library(scales)

# Load CSV files
survey_2017 <- read_csv("data/survey_2017.csv")
survey_2022 <- read_csv("data/survey_2022.csv")
survey <- bind_rows(survey_2017, survey_2022)

# Set Likert Levels
likert_levels <- c(
  "Very difficult", "Difficult", "Somewhat difficult", "Neither",
  "Somewhat manageable", "Manageable", "Very manageable"
)

# Mutate factors and buckets
survey <- survey %>%
  mutate(
    paycheque_delay_difficulty = factor(paycheque_delay_difficulty, levels = likert_levels, ordered = TRUE),
    emergency_fund_likelihood  = factor(emergency_fund_likelihood, levels = likert_levels, ordered = TRUE),
    financial_wellness_cluster = factor(financial_wellness_cluster, levels = c("Stressed", "Coping", "Comfortable")),
    paycheque_delay_score      = as.numeric(paycheque_delay_difficulty),
    emergency_fund_score       = as.numeric(emergency_fund_likelihood),
    age_bucket = cut(age, breaks = c(18, 25, 35, 45, 55, 66), labels = c("18-24", "25-34", "35-44", "45-54", "55-65"), right = FALSE, include.lowest = TRUE),
    income_bucket = cut(income, breaks = c(30000, 50000, 75000, 100000, 150000, Inf), labels = c("$30k-49k", "$50k-74k", "$75k-99k", "$100k-149k","$150k+"), right = FALSE, include.lowest = TRUE),
    province_bucket = case_when(
      province == "Alberta" ~ "AB",
      province == "British Columbia" ~ "BC",
      province == "Ontario" ~ "ON",
      province == "Quebec" ~ "QC",
      TRUE ~ "Other"
    )
  )

survey_2017 <- survey %>% filter(year == 2017)
survey_2022 <- survey %>% filter(year == 2022)


#| fig-width: 9
#| fig-height: 5
## Try replacing paycheque_delay_score with emergency_fund_score or income
ggplot(survey, aes(x = financial_wellness_cluster, y = paycheque_delay_score, fill = factor(year))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Paycheque Delay Difficulty by Cluster: 2017 vs 2022",
       subtitle = "Score: 1 = Very difficult ... 7 = Very manageable",
       x = "Cluster", y = "Difficulty Score", fill = "Year") +
  theme_minimal(base_size = 14)
```

## Next Time:
How to clean the data so that we can work with it?

