---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 4: Visualization - Day 1"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "September 22, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
    - ggrepel
    - patchwork
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
theme_set(theme_gray(base_size = 14))
```

## Where we are

Chapter 1 gave you a working template for plots. This chapter explains
*why* that template works.

By the end you should be able to describe **any** ggplot2 plot in terms of
seven components:

- data
- aesthetic mappings
- a geom
- a stat
- a position adjustment
- a coordinate system
- a faceting scheme

::: {.notes}
Frame this as moving from "recipes" to "grammar". Once students see the seven
slots, unfamiliar plots stop looking like new syntax to memorize.
:::

## Setup

```{r}
#| eval: false
library(tidyverse)
```

We'll work mostly with two datasets that ship with ggplot2:

- `mpg` — 234 car models, 11 variables
- `diamonds` — ~54,000 diamonds

```{r}
mpg |> select(manufacturer, model, displ, hwy, class, drv) |> head(4)
```

::: {.callout-note}
Every dataset in this deck (`mpg`, `diamonds`, `smaller`, `label_info`, ...)
is already loaded in the live-code cells below via a hidden setup chunk — you
don't need to re-run `library(tidyverse)` yourself before trying an exercise.
:::

# Aesthetic mappings

## Mapping a variable to an aesthetic

An **aesthetic** is a visual property of the marks on the plot: position,
color, shape, size, transparency.

`aes()` says "let this variable *control* that property."

```{r}
#| fig-height: 3.4
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Not every aesthetic suits every variable

`shape` only supplies six distinct symbols before it gives up — a seventh
category simply goes unplotted, with a warning.

```{r}
#| fig-height: 3
#| warning: true
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```

::: {.callout-warning}
`class` has seven levels but `shape` can only display six. SUVs (62 rows)
silently disappear from the plot above — always read the warnings ggplot2
gives you, they're telling you data went missing from the picture.
:::

## Ordered aesthetics imply ordered data

`size` and `alpha` are *ranked* channels. Mapping an unordered category to
them tells the reader a ranking exists when it doesn't.

```{r}
#| fig-height: 3
#| warning: true
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()
```

::: {.callout-caution}
`class` (2seater, compact, midsize, ...) has no inherent order. Mapping it to
`size` or `alpha` visually implies "bigger/darker = more," which is
misleading for a purely categorical variable. Save these aesthetics for
variables that are genuinely ordered or continuous.
:::

## Mapping vs. setting

Inside `aes()` = mapping, driven by data, gets a legend.
Outside `aes()` = setting, a fixed constant, no legend.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue", size = 2, shape = 17)
```

## Setting values

When you set an aesthetic you supply a raw value, not a variable:

| Aesthetic | Value |
|---|---|
| `color` | a color name or hex string, `"blue"`, `"#E81B23"` |
| `size`  | point size in mm, `size = 1` |
| `shape` | an integer 0–25, `shape = 1` |
| `alpha` | opacity from 0 to 1 |

Shapes 0–14 are hollow (border from `color`), 15–20 are solid (filled with
`color`), and 21–25 take both a `color` border and a `fill`.

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Draw `hwy` vs. `displ` with pink filled triangles.
2. Why does the code below *not* produce blue points?
3. What does `stroke` do, and which shapes respond to it? (`?geom_point`)
4. What happens with `aes(color = displ < 5)`?
:::

```{r}
#| eval: false
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```

```{webr}
#| exercise: aes_mappings

```

# Geometric objects

## A geom is how the data gets drawn

Same data, same mappings, different geom:

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4.5
#| fig-height: 3.5
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4.5
#| fig-height: 3.5
ggplot(mpg, aes(displ, hwy)) +
  geom_smooth()
```
:::
::::

## Geoms accept different aesthetics

A line has no shape; a point has no linetype. ggplot2 quietly ignores
aesthetics a geom can't use.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) +
  geom_smooth()
```

Mapping a discrete variable splits the geom into one object per level.

## Grouping without a legend

`group` splits the data the same way, but adds no legend and no visual
distinction.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv))
```

::: {.callout-note}
`group` is what mapping a discrete variable does *implicitly* behind the
scenes — using it explicitly is handy when you want the splitting behaviour
without the legend or the visual distinction that `color` or `linetype`
would add.
:::

## Layers stack

Multiple geoms in one plot. Mappings in `ggplot()` are global; mappings in a
geom are local to that layer.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE)
```

## Layers can have their own data

Useful for highlighting a subset:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "grey60") +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red", size = 2
  )
```

## Different geoms reveal different things

```{r}
#| fig-height: 3
#| layout-ncol: 3
#| fig-width: 3.2
ggplot(mpg, aes(x = hwy)) + geom_histogram(binwidth = 2)
ggplot(mpg, aes(x = hwy)) + geom_density()
ggplot(mpg, aes(x = hwy)) + geom_boxplot()
```

The histogram and density show bimodality and right skew; the boxplot
compresses that into a summary but flags outliers.

## Beyond the built-in geoms

ggplot2 ships with 40+ geoms.

::: {.callout-tip}
If you need a geom that doesn't exist, check the extension gallery before
writing your own: <https://exts.ggplot2.tidyverse.org/gallery/>. Someone has
often already solved the problem — e.g. **ggridges** for ridgeline plots.
:::

```{r}
#| eval: false
library(ggridges)

ggplot(mpg, aes(x = hwy, y = drv, fill = drv, color = drv)) +
  geom_density_ridges(alpha = 0.5, show.legend = FALSE)
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Which geom draws a line chart? A boxplot? A histogram? An area chart?
2. What does `show.legend = FALSE` do, and why might you want it?
3. What does `se` control in `geom_smooth()`?
4. Recreate a plot with points colored by `drv` and *one* smooth curve fitted
   to the whole dataset (not one per group).
:::

```{webr}
#| exercise: geoms

```

# Facets

## `facet_wrap()`: one variable

Splits the plot into panels, one per level.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```

## `facet_grid()`: two variables

A two-sided formula, `rows ~ cols`.

```{r}
#| fig-height: 3.4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

Empty panels mean that combination doesn't occur in the data.

## Free scales

By default every panel shares one scale, which is what makes panels
comparable. Release that with `scales`:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl, scales = "free")
```

::: {.callout-note}
`"free_x"`, `"free_y"`, or `"free"` let each row/column pick its own range —
easier to read within a panel, but panels are no longer directly comparable
to each other. Use free scales for exploration, shared scales for
communication.
:::

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What happens if you facet on a continuous variable?
2. What does `.` mean in `facet_grid(drv ~ .)`?
3. Faceting vs. the color aesthetic — when is each better? Does the answer
   change as the dataset grows?
4. Which reads better for comparing `displ` across drive trains: faceting by
   row or by column? What does that suggest as a rule of thumb?
:::

```{webr}
#| exercise: facets

```

# Statistical transformations

## Where does `count` come from?

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

`count` isn't a column in `diamonds`. The bar geom computed it.

## Stats

A **stat** is the algorithm a geom runs on your data before drawing.

- bar charts, histograms, freqpolys → bin and count
- smoothers → fit a model, plot predictions
- boxplots → compute a five-number summary

Every geom has a default stat; every stat has a default geom. `?geom_bar`
tells you it uses `stat_count()`, and lists the variables it computes.

## Overriding the stat

When you already have the counts, switch to the identity stat (or use
`geom_col()`):

```{r}
#| fig-height: 3
diamonds |>
  count(cut) |>
  ggplot(aes(x = cut, y = n)) +
  geom_bar(stat = "identity")
```

## Using a computed variable

`after_stat()` reaches into the variables the stat produced.

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```

::: {.callout-important}
Without `group = 1`, each bar is treated as its own group and every
proportion comes out as 1 — there's nothing to divide by. Try deleting
`group = 1` in the live cell below and see what happens.
:::

## Calling a stat directly

Sometimes the summary *is* the point, and naming the stat says so:

```{r}
#| fig-height: 3
ggplot(diamonds) +
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min, fun.max = max, fun = median
  )
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What geom does `stat_summary()` use by default? Rewrite the previous plot
   with it.
2. How does `geom_col()` differ from `geom_bar()`?
3. List some geom/stat pairs. What do they have in common?
4. Why did the proportion bar chart need `group = 1`? Confirm by removing it
   in the cell below.
:::

```{webr}
#| exercise: stats

ggplot(diamonds, aes(x = cut, y = after_stat(prop))) +
  geom_bar()
```

# Position adjustments

## `color` vs. `fill` for bars

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = drv, color = drv)) + geom_bar()
ggplot(mpg, aes(x = drv, fill = drv)) + geom_bar()
```

## Stacking is a position adjustment

Map `fill` to a second variable and the bars stack automatically:

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar()
```

## The four options

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = drv, fill = class)) + geom_bar(position = "fill")
ggplot(mpg, aes(x = drv, fill = class)) + geom_bar(position = "dodge")
```

- `"stack"` — the default for bars
- `"fill"` — stack, then normalize each bar to the same height: compares proportions
- `"dodge"` — side by side: compares individual counts
- `"identity"` — draw where the values fall; overlaps for bars, default for points

## Overplotting

`hwy` and `displ` are rounded, so points land on a grid and pile up. This
scatterplot shows 126 visible points for 234 observations.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Jittering

`position = "jitter"` adds a little random noise. Less accurate at small
scales, far more honest at large ones.

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter()
```

`geom_jitter()` is shorthand for `geom_point(position = "jitter")`.

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What's wrong with a plain `geom_point()` of `cty` vs. `hwy`? Fix it below.
2. Is there any difference between the default and `position = "identity"`
   for points? Why?
3. Which arguments to `geom_jitter()` control the amount of jitter?
4. Compare `geom_jitter()` and `geom_count()`.
:::

```{webr}
#| exercise: positions

ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
```

# Coordinate systems

## The default is Cartesian

x and y act independently. Two alternatives come up often enough to know:

- `coord_quickmap()` — fixes the aspect ratio for geographic data
- `coord_polar()` — bends the Cartesian plane into polar coordinates

```{r}
#| eval: false
nz <- map_data("nz")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```

## Polar coordinates

A bar chart and a Coxcomb chart are the same plot in different coordinates:

```{r}
#| fig-height: 3
#| layout-ncol: 2
#| fig-width: 4
bar <- ggplot(diamonds) +
  geom_bar(aes(x = clarity, fill = clarity), show.legend = FALSE, width = 1) +
  theme(aspect.ratio = 1)

bar + coord_flip()
bar + coord_polar()
```

# Putting it together

## The extended template

```r
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

Seven slots. In practice you fill in three — data, mappings, geom — because
ggplot2 supplies sensible defaults for the rest.

## Why this matters

Any plot you can imagine is some combination of:

data → stat → geom → aesthetic mappings → coordinate system → position → facets

Change one slot, get a different plot. That's the whole idea behind
"the layered grammar of graphics," and it's why you can build hundreds of
thousands of plots from one template.

The original paper is worth reading:
<https://vita.had.co.nz/papers/layered-grammar.pdf>

## Summary

- **Aesthetics** map variables to visual properties; setting differs from mapping
- **Geoms** decide how marks are drawn; layers stack and can override data and mappings
- **Facets** split one plot into comparable panels
- **Stats** compute the values a geom actually plots
- **Positions** resolve overlap: stack, fill, dodge, jitter
- **Coordinates** redefine what x and y mean

Still to come: **themes**, in Chapter 11.

## Resources

- ggplot2 cheatsheet — <https://posit.co/resources/cheatsheets>
- Package reference — <https://ggplot2.tidyverse.org/reference>
- Extension gallery — <https://exts.ggplot2.tidyverse.org/gallery/>
- *ggplot2: Elegant Graphics for Data Analysis* — <https://ggplot2-book.org/>

::: {.notes}
Closing point worth repeating: when a geom you want doesn't exist, check the
extensions before building it yourself.
:::
