---
title: "DATA100 Introduction to Data Analytics"
subtitle: "Lecture 11: Communication"
author: "Agassi Iu"
institute: "Wilfrid Laurier University"
date: "Date, 2026"
format: 
  live-revealjs:
    theme: sky
    slide-number: true
    transition: slide
    incremental: true
    embed-resources: true # Embeds figures and libraries into a single portable HTML file
filters:
  - live
webr:
  packages:
    - ggplot2
    - dplyr
    - tidyr
    - tibble
    - readr
    - stringr
    - forcats
    - purrr
    - scales
    - ggrepel
    - patchwork
execute:
  echo: true
  warning: false   # Hides warnings and package loading messages
  message: false
  fig-width: 7
  fig-height: 4
  fig-align: center
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)
theme_set(theme_gray(base_size = 14))
```

```{webr}
#| context: setup
#| autorun: true
#| echo: false
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)
theme_set(theme_gray(base_size = 14))

label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)

potential_outliers <- mpg |> filter(hwy > 40 | (hwy > 20 & displ > 5))
```

## Two kinds of plot

**Exploratory.** You know what's in it before you look. You make it, glance
at it, move on. You'll make hundreds and discard most.

**Expository.** Someone else has to read it — someone without your background
and without much patience.

Getting from the first to the second takes real effort. This chapter is the
toolkit.

## Setup

```{r}
#| eval: false
library(tidyverse)
library(scales)     # label and break helpers
library(ggrepel)    # non-overlapping text labels
library(patchwork)  # combining plots
```

Pair this chapter with a book about *what* to show, not just how. Cairo's
*The Truthful Art* is a good companion.

::: {.callout-note}
`label_info` and `potential_outliers` (used later in this deck) are
pre-computed in every live-code cell below, alongside `mpg` and `diamonds`.
:::

# Labels

## `labs()`

```{r}
#| fig-height: 3
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

## What each label is for

| Label | Job |
|---|---|
| `title` | State the **finding**, not the plot's contents |
| `subtitle` | Add detail, in smaller type below the title |
| `caption` | Usually the data source |
| `x`, `y` | Replace variable names with descriptions **and units** |
| aesthetic names | Same, for each legend |

::: {.callout-tip}
"A scatterplot of displacement vs. fuel economy" wastes the most prominent
text on the slide describing what the reader can already see. Use the title
to say what they should *conclude*.
:::

## Mathematical labels

Swap the string for `quote()`; see `?plotmath` for the vocabulary.

```{r}
#| fig-height: 2.9
df <- tibble(x = 1:10, y = cumsum(x^2))

ggplot(df, aes(x, y)) +
  geom_point() +
  labs(
    x = quote(x[i]),
    y = quote(sum(x[i] ^ 2, i == 1, n))
  )
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Build a plot from `mpg` with a custom `title`, `subtitle`, `caption`, `x`,
   `y`, and `color`.
2. Recreate a `hwy` vs. `cty` scatterplot where both color *and* shape vary
   by drive train.
3. Think of a graphic you've made recently — what title would state its
   finding rather than describe its axes?
:::

```{webr}
#| exercise: labels

ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
```

# Annotations

## Labelling observations with `geom_text()`

`geom_text()` is `geom_point()` plus a `label` aesthetic. Build a small data
frame of just the things you want labelled:

```{r}
label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)

label_info
```

## Labels instead of a legend

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_text(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, hjust = "right", vjust = "bottom"
  ) +
  theme(legend.position = "none")
```

`hjust` and `vjust` control alignment relative to the anchor point.

## `ggrepel` fixes the overlap

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info,
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_y = 2
  ) +
  theme(legend.position = "none")
```

## Highlighting specific points

A second layer of large hollow points draws the eye to the labelled cases:

```{r}
#| fig-height: 3
potential_outliers <- mpg |> filter(hwy > 40 | (hwy > 20 & displ > 5))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_text_repel(data = potential_outliers, aes(label = model)) +
  geom_point(data = potential_outliers, color = "red") +
  geom_point(
    data = potential_outliers,
    color = "red", size = 3, shape = "circle open"
  )
```

## Other annotation geoms

::: {.callout-note}
- `geom_hline()` / `geom_vline()` — reference lines. Thick and white
  (`linewidth = 2, color = "white"`), drawn *under* the data, reads clearly
  without competing with it
- `geom_rect()` — box off a region with `xmin`/`xmax`/`ymin`/`ymax`.
  **ggforce**'s `geom_mark_hull()` is a nicer alternative for irregular sets
- `geom_segment(arrow = ...)` — point at something. `x`/`y` start,
  `xend`/`yend` end
:::

## `annotate()` for one-off text

Geoms highlight *subsets of data*; `annotate()` adds *individual elements*
without needing a data frame.

```{r}
trend_text <- "Larger engine sizes tend to have lower fuel economy." |>
  str_wrap(width = 30)

trend_text
```

`str_wrap()` inserts line breaks at a chosen width.

## Text plus arrow

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  annotate(
    geom = "label", x = 3.5, y = 38,
    label = trend_text, hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed")
  )
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Use `geom_text()` with infinite positions to place text in all four
   corners.
2. Use `annotate()` to drop a single customized point in the middle of a plot.
3. How does `geom_text()` interact with faceting? How do you label just one
   facet? A different label in each?
4. Which `geom_label()` arguments control the background box?
5. What are the four arguments to `arrow()`?
:::

```{webr}
#| exercise: annotations

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

# Scales

## Scales are always there

These two produce identical output:

```{r}
#| eval: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_color_discrete()
```

Naming scheme: `scale_` + aesthetic + `_` + scale type
(`continuous`, `discrete`, `date`, `datetime`).

## Two reasons to override

1. **Tweak the default** — different breaks, different labels
2. **Replace it entirely** — a transformation or palette that fits your data
   better than a general-purpose default can

## Breaks and labels

Axes and legends are collectively called **guides**. `breaks` sets where the
ticks or keys go; `labels` sets what they say.

```{r}
#| fig-height: 3
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

## Suppressing and renaming

`labels = NULL` removes them entirely — useful for maps, or when you can't
publish absolute numbers. For discrete scales, pass a named vector.

```{r}
#| fig-height: 2.9
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  scale_color_discrete(labels = c("4" = "4-wheel", "f" = "front", "r" = "rear"))
```

## Formatting with the scales package

```{r}
#| fig-height: 2.9
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(labels = label_dollar())

ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, suffix = "K"),
    breaks = seq(1000, 19000, by = 6000)
  )
```

Note `breaks` stays in the original data scale.

## Percentages

```{r}
#| fig-height: 3.2
ggplot(diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  scale_y_continuous(name = "Percentage", labels = label_percent())
```

## Breaks that mark real events

With few data points, put the ticks where the observations are:

```{r}
#| fig-height: 3
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(name = NULL, breaks = presidential$start, date_labels = "'%y")
```

Dates use `date_labels` (a format string) and `date_breaks` (`"2 days"`,
`"1 month"`).

## Legend position

`legend.position` is a **theme** setting:

```{r}
#| fig-height: 2.7
#| layout-ncol: 2
#| fig-width: 4.5
base <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(aes(color = class))

base + theme(legend.position = "right")
base + theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3))
```

Short and wide → top or bottom. Tall and narrow → left or right.
`"none"` removes it.

## Overriding legend aesthetics

```{r}
#| fig-height: 2.9
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, override.aes = list(size = 4)))
```

::: {.callout-warning}
Legend keys inherit whatever aesthetics the layer uses — including a low
`alpha`. Plot many points at `alpha = 1/20` and the legend swatches become
nearly invisible. `override.aes = list(alpha = 1)` (or `size = ...` as above)
fixes the legend without touching the data layer.
:::

## Transform in the scale, not the mapping

```{r}
#| fig-height: 2.9
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(diamonds, aes(x = log10(carat), y = log10(price))) +
  geom_bin2d()

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10()
```

::: {.callout-note}
Visually identical — but the right-hand axes are labelled in the *original*
units. Transforming inside `aes()` leaves the reader staring at log values;
transforming via the scale keeps the plot interpretable.
:::

## Color: ColorBrewer

The default categorical palette spaces hues evenly around the color wheel.
ColorBrewer's palettes are hand-tuned so common forms of color blindness can
still distinguish them.

```{r}
#| fig-height: 2.7
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(aes(color = drv))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  scale_color_brewer(palette = "Set1")
```

## Redundant encoding is simpler still

::: {.callout-tip}
With only a few categories, map shape *and* color to the same variable. It
survives color blindness, photocopying, and black-and-white printing — cheap
insurance for very little extra code.
:::

```{r}
#| fig-height: 3
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_color_brewer(palette = "Set1")
```

## When the mapping is predetermined

Some variables come with expected colors. `scale_color_manual()` pins them
down:

```{r}
#| fig-height: 3
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))
```

## Continuous color: viridis

Designed to be perceptually uniform *and* legible in grayscale. Available as
continuous (`_c`), discrete (`_d`), and binned (`_b`).

```{r}
#| fig-height: 2.6
#| layout-ncol: 2
#| fig-width: 4
df <- tibble(x = rnorm(10000), y = rnorm(10000))

ggplot(df, aes(x, y)) + geom_hex() + coord_fixed() +
  labs(title = "Default", x = NULL, y = NULL)

ggplot(df, aes(x, y)) + geom_hex() + coord_fixed() +
  scale_fill_viridis_c() +
  labs(title = "Viridis", x = NULL, y = NULL)
```

Every color scale exists as both `scale_color_*()` and `scale_fill_*()`, in
US and UK spellings.

## Three ways to set limits

1. Filter the data
2. Set `limits` on the scale
3. Set `xlim`/`ylim` in `coord_cartesian()`

::: {.callout-caution}
Options 1 and 2 both throw data away before any stat or smooth is fit — a
`geom_smooth()` curve will genuinely change shape. Option 3 is a true zoom:
the smooth is fit on the full data and only the *view* changes. Pick option 3
unless you actually want to refit on a subset.
:::

## Zooming, demonstrated

```{r}
#| fig-height: 2.8
#| layout-ncol: 2
#| fig-width: 4.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) + geom_smooth() +
  scale_x_continuous(limits = c(5, 6)) +
  scale_y_continuous(limits = c(10, 25))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) + geom_smooth() +
  coord_cartesian(xlim = c(5, 6), ylim = c(10, 25))
```

Same window, different smooth curves — the left one refit on a subset.

## Expanding limits to share scales

`limits` is genuinely useful in the other direction: making separate plots
comparable.

```{r}
#| fig-height: 2.7
#| layout-ncol: 2
#| fig-width: 4.5
suv     <- mpg |> filter(class == "suv")
compact <- mpg |> filter(class == "compact")

x_scale   <- scale_x_continuous(limits = range(mpg$displ))
y_scale   <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_color_discrete(limits = unique(mpg$drv))

ggplot(suv, aes(displ, hwy, color = drv)) +
  geom_point() + x_scale + y_scale + col_scale
ggplot(compact, aes(displ, hwy, color = drv)) +
  geom_point() + x_scale + y_scale + col_scale
```

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. Why doesn't `scale_color_gradient()` change anything on a `geom_hex()` plot?
2. What's the first argument to every scale? How does it relate to `labs()`?
3. Improve the presidential terms plot: combine the color and break
   customizations, fix the y axis, label each term with a name, add proper
   labels, and place breaks every four years.
4. Use `override.aes` to make a legend readable when `alpha = 1/20`.
:::

```{webr}
#| exercise: scales_ex

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 1/20)
```

# Themes

## Built-in themes

```{r}
#| fig-height: 3.2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

Eight ship with ggplot2: `theme_gray()` (default), `theme_bw()`,
`theme_light()`, `theme_classic()`, `theme_linedraw()`, `theme_dark()`,
`theme_minimal()`, `theme_void()`. **ggthemes** adds many more.

## Why is the default background gray?

::: {.callout-note}
Deliberate. The white grid lines stay visible — they matter for position
judgments — without shouting. The gray gives the panel a typographic weight
similar to surrounding text, so plots sit in a document instead of leaping
off the page. And the continuous field of color makes the plot read as one
object.
:::

## Customizing individual components

`theme()` controls everything non-data. `element_*()` functions describe how
each piece looks.

```{r}
#| fig-height: 2.8
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(
    title = "Larger engine sizes tend to have lower fuel economy",
    caption = "Source: https://fueleconomy.gov."
  ) +
  theme(
    legend.position = c(0.6, 0.7),
    legend.direction = "horizontal",
    legend.box.background = element_rect(color = "black"),
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0)
  )
```

## Worth knowing

- `element_text()` — font face, size, color, justification
- `element_rect()` — backgrounds and borders
- `element_line()` — grid lines and axis lines
- `element_blank()` — remove a component entirely

`plot.title.position = "plot"` aligns the title to the whole graphic rather
than the panel, which usually looks better with left-aligned captions.

See `?theme` for the full list.

# Layout

## patchwork

Save plots as objects, then combine them with operators:

```{r}
#| fig-height: 2.9
p1 <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) +
  geom_boxplot() + labs(title = "Plot 2")

p1 + p2
```

patchwork doesn't add a new function here — it teaches `+` a new trick.

## Composing layouts

`|` places side by side, `/` starts a new row.

```{r}
#| fig-height: 3.4
p3 <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() + labs(title = "Plot 3")

(p1 | p3) / p2
```

## Collecting legends

```{r}
#| fig-height: 4.2
#| echo: true
p1 <- ggplot(mpg, aes(drv, cty, color = drv)) +
  geom_boxplot(show.legend = FALSE) + labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(drv, hwy, color = drv)) +
  geom_boxplot(show.legend = FALSE) + labs(title = "Plot 2")
p3 <- ggplot(mpg, aes(cty, color = drv, fill = drv)) +
  geom_density(alpha = 0.5) + labs(title = "Plot 3")
p4 <- ggplot(mpg, aes(hwy, color = drv, fill = drv)) +
  geom_density(alpha = 0.5) + labs(title = "Plot 4")
p5 <- ggplot(mpg, aes(cty, hwy, color = drv)) +
  geom_point(show.legend = FALSE) + facet_wrap(~drv) + labs(title = "Plot 5")

(guide_area() / (p1 + p2) / (p3 + p4) / p5) +
  plot_annotation(
    title = "City and highway mileage for cars with different drive trains",
    caption = "Source: https://fueleconomy.gov."
  ) +
  plot_layout(guides = "collect", heights = c(1, 3, 2, 4)) &
  theme(legend.position = "top")
```

## What just happened

- `guide_area()` reserves a slot for the shared legend
- `plot_layout(guides = "collect")` gathers legends from every panel into it
- `heights = c(1, 3, 2, 4)` divides the vertical space proportionally

::: {.callout-note}
`&` applies a theme setting to **every** plot in the patchwork; a plain `+`
would only affect the last one added. This is the one place in ggplot2 code
where `&` and `+` are not interchangeable — worth flagging explicitly.
:::

More at <https://patchwork.data-imaginist.com>

## Your turn {.smaller}

::: {.callout-tip}
## Try it yourself
1. What changes if you drop the parentheses from `(p1 | p2) / p3`? Why?
2. Build a patchwork where Plot 1 spans the top row and Plots 2 and 3 share
   the second row, tagged "Fig. A", "Fig. B", "Fig. C".
:::

```{webr}
#| exercise: layout

p1 <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) +
  geom_boxplot() + labs(title = "Plot 2")
p3 <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() + labs(title = "Plot 3")

p1 | p2 / p3
```

## Summary

- **Labels** — the title states the finding; axes carry units
- **Annotations** — `geom_text()`/`ggrepel` for data, `annotate()` for
  one-offs; reference lines, rectangles, and arrows all help
- **Scales** — `breaks` and `labels` to tweak, whole scales to replace;
  transform in the scale so axes stay readable
- **Color** — ColorBrewer and viridis for accessibility; redundant shape
  encoding is free insurance
- **Zooming** — `coord_cartesian()` zooms, `limits` subsets
- **Themes** — everything that isn't data
- **Layout** — patchwork for multi-panel figures

## Going further

- *ggplot2: Elegant Graphics for Data Analysis* — <https://ggplot2-book.org>
- *R Graphics Cookbook*, Winston Chang — <https://r-graphics.org>
- *Fundamentals of Data Visualization*, Claus Wilke —
  <https://clauswilke.com/dataviz/>
- ColorBrewer — <https://colorbrewer2.org/>
